---
title: "Analytical Mapping"
date: "2026-06-14"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: auto
---

------------------------------------------------------------------------

## 1. Learning Outcome

By the end of this hands-on exercise, I will be able to use appropriate R methods to plot analytical maps. Specifically, I will be able to:

-   import geospatial data in rds format into R,
-   create cartographic quality choropleth maps using **tmap** and **tidyverse** functions,
-   develop rate maps,
-   generate percentile maps, and
-   construct box maps.

## 2. Getting Started

### 2.1 Installing and Loading R Packages

The R packages required for this exercise are: **tmap**, **tidyverse**, and **sf**. The code chunk below uses `pacman::p_load()` to check, install, and load all the necessary packages.

```{r}
pacman::p_load(tmap, tidyverse, sf)
```

### 2.2 Importing Data

For this exercise, I will use *NGA_wp.rds*, which contains polygon feature data on Nigeria water points at the Local Government Area (LGA) level.

```{r}
NGA_wp <- read_rds("Hands-on Exercises/Hands-on W9/NGA_wp.rds")
```

## 3. Basic Choropleth Mapping

### 3.1 Visualising Distribution of Non-Functional Water Points

The code chunk below creates a choropleth map showing the distribution of functional water points by LGA.

```{r}
p1 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "wp_functional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Distribution of functional water point by LGAs")
```

The code chunk below creates a choropleth map showing the distribution of total water points by LGA.

```{r}
p2 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "total_wp",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Distribution of total water point by LGAs")

tmap_arrange(p2, p1, nrow = 1)
```

::: callout-note
Placing the two maps side by side allows for easy comparison between total water points and functional water points per LGA.
:::

::: callout-tip
## Your turn

Plot a choropleth map showing the distribution of non-functional water point by LGA.
:::

```{r}
p3 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "wp_nonfunctional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Distribution of non-functional water point by LGAs")

tmap_arrange(p1, p3, nrow = 1)
```

::: callout-note
Comparing the functional and non-functional maps side by side shows that LGAs with a high count of functional water points do not necessarily have a low count of non-functional ones — some LGAs simply have a larger total number of water points overall, so both counts are high. This is exactly why a rate map (covered next) is more informative than mapping raw counts.
:::

## 4. Choropleth Map for Rates

### 4.1 Deriving Proportion of Functional and Non-Functional Water Points

Mapping absolute counts can be misleading because the number of water points varies spatially across LGAs. A rate map, which shows the proportion of functional or non-functional water points, is more informative.

The code chunk below uses `mutate()` of the **dplyr** package to derive two new variables, `pct_functional` and `pct_nonfunctional`.

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional / total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional / total_wp)
```

### 4.2 Plotting the Rate Map

::: callout-tip
## Your turn

Plot a choropleth map showing the distribution of percentage functional water point by LGA.
:::

The code chunk below is used to plot a choropleth map showing the rate of functional water points.

```{r}
tm_shape(NGA_wp) +
  tm_polygons("pct_functional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Rate map of functional water point by LGAs")
```

::: callout-note
Unlike the absolute count maps in Section 3, this rate map controls for the total number of water points in each LGA. LGAs that appeared "high" in the functional count map earlier are not necessarily the ones with the best functional rate — some simply have more water points overall. The rate map gives a fairer, more meaningful basis for comparing LGAs.
:::

## 5. Extreme Value Maps

Extreme value maps are variations of common choropleth maps where the classification is designed to highlight extreme values at the lower and upper ends of the scale, with the goal of identifying outliers.

### 5.1 Percentile Map

The percentile map uses a special quantile classification scheme with six categories: 0–1%, 1–10%, 10–50%, 50–90%, 90–99%, and 99–100%.

#### 5.1.1 Data Preparation

**Step 1: Exclude records with NA values**

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

**Step 2: Create the customised classification and inspect quantile values**

```{r}
percent <- c(0, .01, .1, .5, .9, .99, 1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[, 1], percent)
```

::: callout-note
When variables are extracted from an sf data frame, the geometry column is included by default. The `st_set_geometry(NULL)` function removes the geometry so that base R functions like `quantile()` can operate on the numeric values directly.
:::

#### 5.1.2 Why Write Functions?

The section on percentile maps requires the same sequence of operations to be performed each time we want to create this type of map for a different variable. Writing functions helps to:

-   give the code a meaningful name,
-   maintain all the logic in a single place, and
-   eliminate repetitive copy-and-paste errors.

Source: [Chapter 19: Functions](https://r4ds.had.co.nz/functions.html#functions) of **R for Data Science**.

#### 5.1.3 Creating the get.var Function

The `get.var()` function is used to extract a field as a vector from an sf data frame.

```{r}
get.var <- function(vname, df) {
  v <- df[vname] %>%
    st_set_geometry(NULL)
  v <- unname(v[, 1])
  return(v)
}
```

::: callout-note
-   `vname` is the variable name as a character string.
-   `df` is the sf data frame.
-   The function returns the values as a plain vector without names.
:::

#### 5.1.4 A Percentile Mapping Function

The code chunk below defines the `percentmap()` function, which builds a percentile choropleth map for any variable in the data frame.

```{r}
percentmap <- function(vnam, df, legtitle = NA, mtitle = "Percentile Map") {
  percent <- c(0, .01, .1, .5, .9, .99, 1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)
  tm_shape(df) +
    tm_polygons() +
    tm_shape(df) +
    tm_polygons(vnam,
                title = legtitle,
                breaks = bperc,
                palette = "Blues",
                labels = c("< 1%", "1% - 10%", "10% - 50%",
                           "50% - 90%", "90% - 99%", "> 99%")) +
    tm_borders() +
    tm_layout(main.title = mtitle,
              title.position = c("right", "bottom"))
}
```

#### 5.1.5 Testing the Percentile Mapping Function

```{r}
percentmap("total_wp", NGA_wp)
```

### 5.2 Box Map

A box map is an augmented quartile map with an additional lower and upper category. When there are lower outliers, the map breaks the first class into two: a lower outlier class and a below Q1 class. When there are upper outliers, the map breaks the last class into two: an above Q3 class and an upper outlier class.

The code chunk below is used to create a boxplot to explore the distribution of the `wp_nonfunctional` variable.

```{r}
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```

#### 5.2.1 Creating the boxbreaks Function

The `boxbreaks()` function is used to compute the break points for the box map.

```{r}
boxbreaks <- function(v, mult = 1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  bb <- vector(mode = "numeric", length = 7)
  if (lofence < qv[1]) {
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) {
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

::: callout-note
-   `v` is a vector of observations.
-   `mult` is the multiplier for the IQR, which defaults to 1.5.
-   The function returns a vector of 7 breakpoints that account for the presence or absence of lower and upper outliers.
:::

#### 5.2.2 Creating the get.var Function

The same `get.var()` function defined earlier is used here again to extract variable values from the sf data frame.

```{r}
get.var <- function(vname, df) {
  v <- df[vname] %>%
    st_set_geometry(NULL)
  v <- unname(v[, 1])
  return(v)
}
```

#### 5.2.3 Testing the boxbreaks Function

```{r}
var <- get.var("wp_nonfunctional", NGA_wp)
boxbreaks(var)
```

#### 5.2.4 Boxmap Function

The `boxmap()` function creates the final box map visualisation using the break points computed by `boxbreaks()`.

```{r}
boxmap <- function(vnam, df,
                   legtitle = NA,
                   mtitle = "Box Map",
                   mult = 1.5) {
  var <- get.var(vnam, df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
    tm_shape(df) +
    tm_fill(vnam,
            title = legtitle,
            breaks = bb,
            palette = "Blues",
            labels = c("lower outlier",
                       "< 25%",
                       "25% - 50%",
                       "50% - 75%",
                       "> 75%",
                       "upper outlier")) +
    tm_borders() +
    tm_layout(main.title = mtitle,
              title.position = c("left", "top"))
}

tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```

::: callout-note
-   `vnam` is the variable name (character string).
-   `df` is the simple features polygon layer.
-   `legtitle` is the legend title.
-   `mtitle` is the map title.
-   `mult` is the multiplier for the IQR (default 1.5).
:::

## 6. Reference

-   Kam, T.S. (2023). [*Chapter 23: Analytical Mapping*](https://r4va.netlify.app/chap23). R for Visual Analytics.
