---
title: "A Layered Grammar of Graphics: ggplot2 methods"
date: "2026-04-17"
date-modified: "2026-04-17"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    code-fold: show
    code-summary: "Show the code"
    code-copy: true
editor: visual
---

## Learning Outcome

In this chapter, you will learn the basic ideas behind `ggplot2` and the main components used to build statistical graphics. You will also practise using these components to create data visualisations based on the layered grammar of graphics.

By the end of this chapter, you should be able to use the essential graphical elements in `ggplot2` to produce clear and functional statistical graphics.

## Getting started

::: panel-tabset
## Installing and loading libraries

The code chunk below uses `p_load()` of `pacman` package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse,
               patchwork)
```

## Importing Data

The code chunk below imports `exam_data.csv` into R environment by using `read_csv()` function of `readr` package.

`readr` is one of the tidyverse package.

```{r}
exam_data <- read_csv("Hands-on Exercises/Hands-on W1/Exam_data.csv")
```

## Understanding the Data

```{r}
summary(exam_data)
```

Year end examination grades of a cohort of primary 3 students from a local school.

There are a total of seven attributes. Four of them are categorical data type and the other three are in continuous data type.

Categorical attributes are ID, CLASS, GENDER and RACE.

Continuous attributes are MATHS, ENGLISH and SCIENCE.
:::
