---
title: "Visualising and Analysing Time-oriented Data"
date: "2026-06-08"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

------------------------------------------------------------------------

## 1. Learning Outcome

By the end of this hands-on exercise I will be able to create the following data visualisations by using R packages:

-   plotting a calendar heatmap by using **ggplot2** functions,
-   plotting a cycle plot by using **ggplot2** function,
-   plotting a slopegraph, and
-   plotting a horizon chart.

## 2. Getting Started

### 2.1 Installing and Loading R Packages

Write a code chunk to check, install and launch the following R packages: **scales**, **viridis**, **lubridate**, **ggthemes**, **gridExtra**, **readxl**, **knitr**, **data.table**, **CGPfunctions**, **ggHoriPlot**, and **tidyverse**.

```{r}
pacman::p_load(scales, viridis, lubridate, ggthemes,
               gridExtra, readxl, knitr, data.table,
               ggHoriPlot, tidyverse)
```

## 3. Plotting Calendar Heatmap

In this section, I will learn how to plot a calendar heatmap programmatically by using the **ggplot2** package.

By the end of this section, I will be able to:

-   plot a calendar heatmap by using **ggplot2** functions and extension,
-   write a function using R programming,
-   derive specific date and time related fields by using base R and **lubridate** packages, and
-   perform data preparation tasks by using **tidyr** and **dplyr** packages.

### 3.1 The Data

For the purpose of this hands-on exercise, *eventlog.csv* will be used. This data file consists of 199,999 rows of time-series cyber attack records by country.

### 3.2 Importing the Data

First, I will use the code chunk below to import `eventlog.csv` into R environment and call the data frame as `attacks`.

```{r}
attacks <- read_csv("Hands-on Exercises/Hands-on W8/eventlog.csv")
```

### 3.3 Examining the Data Structure

It is always a good practice to examine the imported data frame before further analysis is performed.

For example, `kable()` can be used to review the structure of the imported data frame.

```{r}
kable(head(attacks))
```

::: callout-note
There are three columns, namely `timestamp`, `source_country` and `tz`.

-   `timestamp` field stores date-time values in POSIXct format.
-   `source_country` field stores the source of the attack. It is in ISO 3166-1 alpha-2 country code.
-   `tz` field stores the time zone of the source IP address.
:::

### 3.4 Data Preparation

**Step 1: Deriving weekday and hour of day fields**

Before we can plot the calendar heatmap, two new fields namely `wkday` and `hour` need to be derived. In this step, I will write a function to perform the task.

```{r}
make_hr_wkday <- function(ts, sc, tz) {
  real_times <- ymd_hms(ts,
                        tz = tz[1],
                        quiet = TRUE)
  dt <- data.table(source_country = sc,
                   wkday = weekdays(real_times),
                   hour = hour(real_times))
  return(dt)
}
```

::: callout-note
-   `ymd_hms()` and `hour()` are from the **lubridate** package, and
-   `weekdays()` is a base R function.
:::

**Step 2: Deriving the attacks tibble data frame**

```{r}
wkday_levels <- c('Saturday', 'Friday',
                  'Thursday', 'Wednesday',
                  'Tuesday', 'Monday',
                  'Sunday')

attacks <- attacks %>%
  group_by(tz) %>%
  do(make_hr_wkday(.$timestamp,
                   .$source_country,
                   .$tz)) %>%
  ungroup() %>%
  mutate(wkday = factor(
    wkday, levels = wkday_levels),
    hour  = factor(
      hour, levels = 0:23))
```

::: callout-note
Beside extracting the necessary data into the `attacks` data frame, `mutate()` of the **dplyr** package is used to convert `wkday` and `hour` fields into factor so they'll be ordered when plotting.
:::

The table below shows the tidy tibble table after processing.

```{r}
kable(head(attacks))
```

### 3.5 Building the Calendar Heatmap

```{r}
grouped <- attacks %>%
  count(wkday, hour) %>%
  ungroup() %>%
  na.omit()

ggplot(grouped,
       aes(hour,
           wkday,
           fill = n)) +
  geom_tile(color = "white",
            size = 0.1) +
  theme_tufte(base_family = "Helvetica") +
  coord_equal() +
  scale_fill_gradient(name = "# of attacks",
                      low = "sky blue",
                      high = "dark blue") +
  labs(x = NULL,
       y = NULL,
       title = "Attacks by weekday and time of day") +
  theme(axis.ticks = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6))
```

::: callout-note
## Things to learn from the code chunk

-   A tibble data table called `grouped` is derived by aggregating the attack by `wkday` and `hour` fields.
-   A new field called `n` is derived by using `group_by()` and `count()` functions.
-   `na.omit()` is used to exclude missing values.
-   `geom_tile()` is used to plot tiles (grids) at each x and y position. The `color` and `size` arguments are used to specify the border colour and line size of the tiles.
-   `theme_tufte()` of the **ggthemes** package is used to remove unnecessary chart junk. To learn which visual components of default ggplot2 have been excluded, you are encouraged to comment out this line to examine the default plot.
-   `coord_equal()` is used to ensure the plot will have an aspect ratio of 1:1.
-   `scale_fill_gradient()` function is used to create a two colour gradient (low–high).
:::

Then I can simply group the count by `hour` and `wkday` and plot it. Since there are values for every combination, there is no need to further preprocess the data.

### 3.6 Building Multiple Calendar Heatmaps

**Challenge: Building multiple heatmaps for the top four countries with the highest number of attacks.**

**Step 1: Deriving attack by country object**

In order to identify the top 4 countries with the highest number of attacks, I am required to:

-   count the number of attacks by country,
-   calculate the percent of attacks by country, and
-   save the results in a tibble data frame.

```{r}
attacks_by_country <- count(
  attacks, source_country) %>%
  mutate(percent = percent(n/sum(n))) %>%
  arrange(desc(n))
```

**Step 2: Preparing the tidy data frame**

In this step, I will extract the attack records of the top 4 countries from the `attacks` data frame and save the data in a new tibble data frame called `top4_attacks`.

```{r}
top4 <- attacks_by_country$source_country[1:4]
top4_attacks <- attacks %>%
  filter(source_country %in% top4) %>%
  count(source_country, wkday, hour) %>%
  ungroup() %>%
  mutate(source_country = factor(
    source_country, levels = top4)) %>%
  na.omit()
```

**Step 3: Plotting the Multiple Calendar Heatmap by using ggplot2 package.**

```{r}
ggplot(top4_attacks,
       aes(hour,
           wkday,
           fill = n)) +
  geom_tile(color = "white",
            size = 0.1) +
  theme_tufte(base_family = "Helvetica") +
  coord_equal() +
  scale_fill_gradient(name = "# of attacks",
                      low = "sky blue",
                      high = "dark blue") +
  facet_wrap(~source_country, ncol = 2) +
  labs(x = NULL, y = NULL,
       title = "Attacks on top 4 countries by weekday and time of day") +
  theme(axis.ticks = element_blank(),
        axis.text.x = element_text(size = 7),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6))
```

## 4. Plotting Cycle Plot

In this section, I will learn how to plot a cycle plot showing the time-series patterns and trend of visitor arrivals from Vietnam programmatically by using **ggplot2** functions.

### 4.1 Step 1: Data Import

For the purpose of this hands-on exercise, *arrivals_by_air.xlsx* will be used.

The code chunk below imports `arrivals_by_air.xlsx` by using `read_excel()` of the **readxl** package and saves it as a tibble data frame called `air`.

```{r}
air <- read_excel("Hands-on Exercises/Hands-on W8/arrivals_by_air.xlsx")
```

### 4.2 Step 2: Deriving Month and Year Fields

Next, two new fields called `month` and `year` are derived from the `Month-Year` field.

```{r}
air$month <- factor(month(air$`Month-Year`),
                    levels=1:12,
                    labels=month.abb,
                    ordered=TRUE)
air$year <- year(ymd(air$`Month-Year`))
```

### 4.3 Step 4: Extracting the Target Country

Next, the code chunk below is used to extract data for the target country (i.e. Vietnam).

```{r}
Vietnam <- air %>%
  select(`Vietnam`,
         month,
         year) %>%
  filter(year >= 2010)
```

### 4.4 Step 5: Computing Year Average Arrivals by Month

The code chunk below uses `group_by()` and `summarise()` of **dplyr** to compute year average arrivals by month.

```{r}
hline.data <- Vietnam %>%
  group_by(month) %>%
  summarise(avgvalue = mean(`Vietnam`))
```

### 4.5 Step 6: Plotting the Cycle Plot

The code chunk below is used to plot the cycle plot.

```{r}
ggplot() +
  geom_line(data=Vietnam,
            aes(x=year,
                y=`Vietnam`,
                group=month),
            colour="black") +
  geom_hline(aes(yintercept=avgvalue),
             data=hline.data,
             linetype=6,
             colour="red",
             size=0.5) +
  facet_grid(~month) +
  labs(axis.text.x = element_blank(),
       title = "Visitor arrivals from Vietnam by air, Jan 2010-Dec 2019") +
  xlab("") +
  ylab("No. of Visitors") +
  theme_tufte(base_family = "Helvetica")
```

## 5. Plotting Slopegraph

In this section I will learn how to plot a slopegraph by using R.

Before getting started, I will note that the original chapter uses `newggslopegraph()` from the **CGPfunctions** package. Since that package is no longer available on CRAN, I will build the slopegraph using **ggplot2** directly to achieve the same result.

### 5.1 Step 1: Data Import

Import the rice data set into R environment by using the code chunk below.

```{r}
rice <- read_csv("Hands-on Exercises/Hands-on W8/rice.csv")
```

### 5.2 Step 2: Plotting the Slopegraph

The code chunk below is used to plot a basic slopegraph comparing rice yields across top Asian countries between 1961 and 1980.

```{r}
rice_slope <- rice %>%
  filter(Year %in% c(1961, 1980)) %>%
  mutate(Year = factor(Year))

ggplot(data = rice_slope,
       aes(x = Year,
           y = Yield,
           group = Country)) +
  geom_line(aes(color = Country),
            linewidth = 0.75,
            alpha = 0.8) +
  geom_point(aes(color = Country),
             size = 2.5) +
  geom_text(data = rice_slope %>% filter(Year == 1961),
            aes(label = Country, color = Country),
            hjust = 1.15,
            size = 3) +
  geom_text(data = rice_slope %>% filter(Year == 1980),
            aes(label = Country, color = Country),
            hjust = -0.15,
            size = 3) +
  scale_x_discrete(expand = expansion(mult = 0.35)) +
  theme_tufte(base_family = "Helvetica") +
  theme(legend.position = "none",
        axis.ticks = element_blank()) +
  labs(title = "Rice Yield of Top 11 Asian Countries",
       subtitle = "1961-1980",
       caption = "Prepared by: Do Thi Tue Minh",
       x = NULL,
       y = "Yield (tonnes/ha)")
```

::: callout-note
## Thing to learn from the code chunk above

For effective data visualisation design, `factor()` is used to convert the value type of the `Year` field from numeric to factor.
:::

## 6. Reference

-   Kam, T.S. (2023). [*Chapter 17: Visualising and Analysing Time-oriented Data*](https://r4va.netlify.app/chap17). R for Visual Analytics.
