---
title: "Choropleth Mapping with R"
date: "2026-06-14"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: auto
---

------------------------------------------------------------------------

## 1. Learning Outcome

By the end of this hands-on exercise, I will be able to:

-   import geospatial and attribute data into R using appropriate R methods,
-   convert imported data into appropriate simple feature data frames,
-   perform a relational join to merge both data sets,
-   create choropleth maps using the **tmap** package, and
-   apply different data classification methods and colour schemes to enhance choropleth maps.

## 2. Getting Started

### 2.1 Installing and Loading R Packages

The R packages required for this exercise are: **sf**, **tmap**, and **tidyverse**. The code chunk below uses `pacman::p_load()` to check, install, and load all the necessary packages.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## 3. Importing Data into R

### 3.1 The Data

Two data sets will be used in this exercise:

-   **Master Plan 2014 Subzone Boundary (Web)** in ESRI shapefile format, which provides the geospatial information of Singapore's planning subzones.
-   **Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2000–2020** in csv format (*respopagesextod2011to2020.csv*), which provides the attribute data on population.

### 3.2 Importing Geospatial Data into R

The code chunk below uses `st_read()` of the **sf** package to import the shapefile into R as a simple feature data frame called `mpsz`.

```{r}
mpsz <- st_read(dsn = "Hands-on Exercises/Hands-on W9/MP14_SUBZONE_WEB_PL",
                layer = "MP14_SUBZONE_WEB_PL")
```

I can examine the content of `mpsz` by using the code chunk below.

```{r}
mpsz
```

::: callout-note
The output reveals that there are a total of 323 multipolygon features and 15 fields in the `mpsz` simple feature data frame. `mpsz` is in **svy21** projected coordinates system with two extent values: xmin, ymin, xmax, ymax.

Notice that only the first ten records will be displayed. Do you know why?

This is because `mpsz` is also a tibble data frame, and by default, R only prints the first 10 rows of a tibble to keep the console output readable. The same truncation behaviour will be seen whenever a large tibble or sf object is printed without `head()` or explicit row limits.
:::

### 3.3 Importing Attribute Data into R

Next, I will import *respopagesextod2011to2020.csv* into RStudio as a tibble data frame called `popdata`.

```{r}
popdata <- read_csv("Hands-on Exercises/Hands-on W9/respopagesextod2011to2020.csv")
```

### 3.4 Data Preparation

Before a choropleth map can be prepared, a data table with values at the planning subzone level is required. The data table should include the variables `PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL`, and `DEPENDENCY`.

-   `YOUNG`: age group 0 to 4 until age group 20 to 24,
-   `ECONOMY ACTIVE`: age group 25 to 29 until age group 60 to 64,
-   `AGED`: age group 65 and above,
-   `TOTAL`: all age groups, and
-   `DEPENDENCY`: ratio of (`YOUNG` + `AGED`) / `ECONOMY ACTIVE`.

#### 3.4.1 Data Wrangling

The following data wrangling and transformation functions will be used:

-   `pivot_wider()` of the **tidyr** package, and
-   `mutate()`, `filter()`, `group_by()`, and `select()` of the **dplyr** package.

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from = AG,
              values_from = POP) %>%
  mutate(YOUNG = rowSums(.[3:6]) + rowSums(.[12])) %>%
  mutate(`ECONOMY ACTIVE` = rowSums(.[7:11]) + rowSums(.[13:15])) %>%
  mutate(`AGED` = rowSums(.[16:21])) %>%
  mutate(`TOTAL` = rowSums(.[3:21])) %>%
  mutate(`DEPENDENCY` = (`YOUNG` + `AGED`) / `ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL`, `DEPENDENCY`)
```

#### 3.4.2 Joining Attribute and Geospatial Data

Before performing the georelational join, I need to convert the values of the `PA` and `SZ` fields to uppercase so they match the values in the geospatial data.

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, `left_join()` is used to merge the geographical data and the attribute table using the planning subzone name as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

```{r}
write_rds(mpsz_pop2020, "Hands-on Exercises/Hands-on W9/mpszpop2020.rds")
```

::: callout-note
`left_join()` of the **dplyr** package is used with `mpsz` as the left data table to ensure that all the planning subzones in Singapore are retained.
:::

## 4. Choropleth Mapping Using tmap

Two approaches can be used to prepare a choropleth map using **tmap**:

-   Plotting a thematic map quickly by using `qtm()`, and
-   Plotting a highly customisable thematic map by using **tmap** elements.

### 4.1 Plotting a Choropleth Map Quickly Using qtm()

The easiest and quickest way to draw a choropleth map using **tmap** is `qtm()`. It is concise and provides a good default visualisation in many cases.

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, fill = "DEPENDENCY")
```

::: callout-note
-   `tmap_mode()` with the "plot" option is used to produce a static map. For an interactive mode, the "view" option should be used.
-   The `fill` argument is used to map the attribute `DEPENDENCY`.
:::

### 4.2 Creating a Choropleth Map Using tmap Elements

Despite its usefulness for drawing a choropleth map quickly and easily, the drawback of `qtm()` is that it makes aesthetic control of individual layers harder. To draw a high-quality cartographic choropleth map, **tmap**'s drawing elements should be used.

#### 4.2.1 Drawing a Base Map

The basic building block of **tmap** is `tm_shape()` followed by one or more layer elements such as `tm_fill()` and `tm_polygons()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

#### 4.2.2 Drawing a Choropleth Map Using tm_polygons()

To draw a choropleth map showing the geographical distribution of a selected variable by planning subzone, I just need to assign the target variable such as `DEPENDENCY` to `tm_polygons()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY")
```

::: callout-note
-   The default interval binning used to draw the choropleth map is called "pretty".
-   The default colour scheme used is `YlOrRd` of ColorBrewer.
-   By default, missing values will be shaded in grey.
:::

#### 4.2.3 Drawing a Choropleth Map Using tm_fill() and tm_borders()

`tm_polygons()` is a wrapper of `tm_fill()` and `tm_borders()`. `tm_fill()` shades the polygons using the default colour scheme, while `tm_borders()` adds the borders of the shapefile onto the choropleth map.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY") +
  tm_borders(lwd = 0.01, fill_alpha = 0.1)
```

::: callout-note
The `alpha` argument is used to define transparency as a number between 0 (totally transparent) and 1 (not transparent). By default, the `alpha` value of the colour is used (normally 1).
:::

### 4.3 Data Classification Methods of tmap

Most choropleth maps employ some method of data classification. The point of classification is to take a large number of observations and group them into data ranges or classes.

**tmap** provides ten data classification methods, namely: *fixed*, *sd*, *equal*, *pretty* (default), *quantile*, *kmeans*, *hclust*, *bclust*, *fisher*, and *jenks*.

::: callout-warning
## Warning: Maps Lie!

The choice of classification method and number of classes can dramatically change the message a choropleth map conveys, even though the underlying data has not changed at all. The same variable can be made to look almost uniform or highly polarised purely by adjusting these two settings. As a map maker, I need to be deliberate and transparent about the classification method and number of classes I choose, since these decisions can unintentionally (or intentionally) mislead readers.
:::

#### 4.3.1 Plotting Choropleth Maps with Built-in Classification Methods

The code chunk below shows a quantile data classification that uses 5 classes.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

The code chunk below shows an equal data classification that uses 5 classes.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

The code chunk below shows a Jenks natural breaks classification that uses 5 classes.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "jenks",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

#### 4.3.2 Plotting Choropleth Maps with Custom Break Points

Besides the built-in data classification methods, breakpoints can also be set explicitly by means of the `breaks` argument. It is good practice to get some descriptive statistics on the variable before setting the breakpoints.

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

With reference to the summary statistics above, I will set break points at 0.60, 0.70, 0.80, and 0.90. In addition, I need to include a minimum and maximum, which I will set at 0 and 1.00. The `breaks` vector is thus `c(0, 0.60, 0.70, 0.80, 0.90, 1.00)`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(fill_alpha = 0.5)
```

#### 4.3.3 DIY: Comparing Different Classification Methods

::: callout-tip
## DIY

Using what I had learned, prepare choropleth maps by using different classification methods supported by **tmap** and compare their differences.
:::

To answer this, I will plot maps using five different classification methods—jenks, quantile, equal, sd, and fisher—all keeping `n = 5`, side by side for comparison.

```{r}
map_jenks <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "jenks", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Jenks")

map_quantile <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile")

map_equal <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "equal", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal")

map_sd <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "sd", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Standard Deviation")

map_fisher <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "fisher", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Fisher")

tmap_arrange(map_jenks, map_quantile, map_equal, map_sd, map_fisher, ncol = 3)
```

::: callout-note
## My observations

-   **Jenks** and **fisher** produce very similar maps because both methods minimise within-class variance and tend to find natural breaks/clusters in the data — subzones with extreme dependency ratios stand out clearly as their own class.
-   **Quantile** forces an equal number of subzones (roughly 1/5 of the total) into each class, so the map looks the most "balanced" even though the underlying values within each class can vary widely. This can hide the fact that most subzones actually have similar dependency ratios.
-   **Equal** divides the value range into 5 equally-sized intervals regardless of how the data points are distributed. Because `DEPENDENCY` is right-skewed, most subzones fall into just one or two of the lower classes, making the map look almost monochrome.
-   **Standard deviation** centres the classes on the mean, so it is most useful for showing which subzones are above or below average, but it is harder to interpret for readers unfamiliar with the method.
-   Overall, the same data can tell quite different visual stories depending on the classification method chosen — this is exactly the risk highlighted in the "Maps Lie!" warning above.
:::

#### 4.3.4 DIY: Comparing Different Numbers of Classes

::: callout-tip
## DIY

Prepare choropleth maps by using a similar classification method but with different numbers of classes (i.e. 2, 6, 10, 20). Compare the output maps. What observation can you draw?
:::

I will hold the classification method constant at "pretty" and vary the number of classes (`n`) to 2, 6, 10, and 20.

```{r}
map_n2 <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "pretty", n = 2)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 2")

map_n6 <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "pretty", n = 6)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 6")

map_n10 <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "pretty", n = 10)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 10")

map_n20 <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "pretty", n = 20)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 20")

tmap_arrange(map_n2, map_n6, map_n10, map_n20, ncol = 2)
```

::: callout-note
## My observations

-   With **n = 2**, the map only distinguishes "high" versus "low" dependency subzones. This is the most generalised view and hides almost all spatial nuance — many genuinely different subzones get lumped into the same broad class.
-   As **n increases to 6 and 10**, finer spatial patterns start to emerge — clusters of planning subzones with similarly high or low dependency ratios become visible, and the map begins to reveal local variation that the 2-class version completely missed.
-   At **n = 20**, the map becomes very granular. While this preserves more detail, it also becomes harder to read at a glance: many classes share visually similar colours, the legend becomes long and unwieldy, and individual classes may contain very few subzones (or none at all), reducing the map's communicative value.
-   The overall takeaway is that there is a trade-off between **detail** and **readability**: too few classes oversimplify and can mask meaningful variation, while too many classes overwhelm the reader and dilute the visual signal. A moderate number of classes (typically 5–7) tends to strike the best balance for most choropleth maps.
:::

### 4.4 Colour Scheme

**tmap** supports colour ramps either defined by the user or a set of predefined colour ramps from the **RColorBrewer** package.

#### 4.4.1 Using ColorBrewer Palette

To change the colour, I assign the preferred colour scheme to the `values` argument of `tm_scale_intervals()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

To reverse the colour shading, I add a "-" prefix to the colour scheme name.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

### 4.5 Map Layouts

Map layout refers to the combination of all map elements into a cohesive map. Map elements include objects to be mapped, the title, the scale bar, the compass, margins, and aspect ratios.

#### 4.5.1 Map Legend

In **tmap**, several legend options are provided to change the placement, format, and appearance of the legend.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "jenks",
                n = 5,
                values = "brewer.greens"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Dependency Ratio by planning subzone\n(Jenks classification)")
```

#### 4.5.2 Map Style

**tmap** allows a wide variety of layout settings to be changed. They can be called by using `tmap_style()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

#### 4.5.3 Cartographic Furniture

**tmap** also provides arguments to draw other map furniture such as compass, scale bar, and grid lines.

In the code chunk below, `tm_compass()`, `tm_grid()`, and `tm_credits()` are used to add a compass rose, grid lines, and attribution text onto the choropleth map.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_grid(alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\nand Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```

To reset the default style, the code chunk below is used.

```{r}
tmap_style("white")
```

### 4.6 Small Multiple Choropleth Maps

**Small multiple maps**, also referred to as **facet maps**, are composed of many maps arranged side by side, and sometimes stacked vertically. Small multiple maps enable the visualisation of how spatial relationships change with respect to another variable, such as time.

In **tmap**, small multiple maps can be plotted in three ways:

-   by assigning multiple values to at least one of the aesthetic arguments,
-   by defining a group-by variable in `tm_facets()`, and
-   by creating multiple standalone maps with `tmap_arrange()`.

#### 4.6.1 By Assigning Multiple Values to at Least One Aesthetic Argument

In this example, small multiple choropleth maps are created by assigning multiple values to `tm_fill()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill(c("YOUNG", "AGED"),
          style = "equal",
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

In this example, small multiple choropleth maps are created by assigning multiple values along with different classification methods.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(c("DEPENDENCY", "AGED"),
              style = c("equal", "quantile"),
              palette = list("Blues", "Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

#### 4.6.2 By Defining a Group-by Variable in tm_facets()

In this example, multiple small maps are created using `tm_facets()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) +
  tm_facets(by = "REGION_N",
            free.coords = TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"),
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

#### 4.6.3 By Creating Multiple Standalone Maps with tmap_arrange()

In this example, multiple small choropleth maps are created by using `tmap_arrange()`.

```{r}
youngmap <- tm_shape(mpsz_pop2020) +
  tm_polygons("YOUNG",
              style = "quantile",
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020) +
  tm_polygons("AGED",
              style = "quantile",
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp = 1, ncol = 2)
```

### 4.7 Mapping Spatial Objects Meeting Selection Criteria

Instead of creating small multiple choropleth maps, the selection function can be used to map spatial objects that meet a specified selection criterion.

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N == "CENTRAL REGION", ]) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          legend.hist = TRUE,
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45,
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

## 5. Reference

-   Kam, T.S. (2023). [*Chapter 21: Choropleth Mapping with R*](https://r4va.netlify.app/chap21). R for Visual Analytics.

### **5.1 All about tmap package**

-   [tmap: Thematic Maps in R](https://www.jstatsoft.org/article/view/v084i06)

-   [tmap](https://cran.r-project.org/web/packages/tmap/index.html)

-   [tmap: get started!](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)

-   [tmap: changes in version 2.0](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-changes-v2.html)

-   [tmap: creating thematic maps in a flexible way (useR!2015)](http://von-tijn.nl/tijn/research/presentations/tmap_user2015.pdf)

-   [Exploring and presenting maps with tmap (useR!2017)](http://von-tijn.nl/tijn/research/presentations/tmap_user2017.pdf)

### **5.2 Geospatial data wrangling**

-   [sf: Simple Features for R](https://cran.r-project.org/web/packages/sf/index.html)

-   [Simple Features for R: StandardizedSupport for Spatial Vector Data](https://journal.r-project.org/archive/2018/RJ-2018-009/RJ-2018-009.pdf)

-   [Reading, Writing and Converting Simple Features](https://cran.r-project.org/web/packages/sf/vignettes/sf2.html)

### **5.3 Data wrangling**

-   [dplyr](https://dplyr.tidyverse.org/)

-   [Tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)

-   [tidyr: Easily Tidy Data with ‘spread()’ and ‘gather()’ Functions](https://cran.r-project.org/web/packages/tidyr/tidyr.pdf)
