---
title: "Funnel Plots for Fair Comparisons"
date: "May 3, 2026"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## 1. Overview

Funnel plot is a specially designed data visualisation for conducting unbiased comparison between outlets, stores or business entities. In this exercise, I will:

-   plot funnel plots by using **funnelPlotR** package,
-   plot static funnel plot by using **ggplot2** package, and
-   plot interactive funnel plot by using both **plotly R** and **ggplot2** packages.

## 2. Installing and Launching R Packages

In this exercise, the following R packages will be used:

-   **readr** for importing csv into R.
-   **FunnelPlotR** for creating funnel plot.
-   **ggplot2** for creating funnel plot manually.
-   **knitr** for building static html table.
-   **plotly** for creating interactive funnel plot.

```{r}
pacman::p_load(tidyverse, FunnelPlotR, plotly, knitr)
```

## 3. Importing Data

For this exercise, I am using the COVID-19 DKI Jakarta dataset downloaded from the [Open Data Covid-19 Provinsi DKI Jakarta portal](https://riwayat-file-covid-19-dki-jakarta-jakartagis.hub.arcgis.com/). The goal is to compare cumulative COVID-19 cases and deaths by sub-district (kelurahan) as at 31st July 2021.

```{r}
covid19 <- read_csv("Hands-on Exercises/Hands-on W4/COVID-19_DKI_Jakarta.csv") %>%
  mutate_if(is.character, as.factor)
```

A preview of the data:

```{r}
kable(head(covid19))
```

## 4. FunnelPlotR Methods

[**FunnelPlotR**](https://nhs-r-community.github.io/FunnelPlotR/) package uses ggplot to generate funnel plots. It requires a `numerator` (events of interest), `denominator` (population to be considered) and `group`. Key arguments I will use:

-   `limit`: plot limits (95 or 99).
-   `label_outliers`: to label outliers (true or false).
-   `Poisson_limits`: to add Poisson limits to the plot.
-   `OD_adjust`: to add overdispersed limits to the plot.
-   `xrange` and `yrange`: to specify the range to display for axes, acts like a zoom function.

### 4.1 The Basic Plot

```{r}
funnel_plot(
  .data = covid19,
  numerator = Positive,
  denominator = Death,
  group = `Sub-district`
)
```

::: callout-note
-   `group` here defines the level of the points to be plotted i.e. Sub-district, District or City. If City is chosen, there are only six data points.
-   By default, `data_type` argument is `"SR"`.
-   `limit`: Plot limits, accepted values are 95 or 99, corresponding to 95% or 99.8% quantiles of the distribution.
:::

### 4.2 Makeover 1

Here I change `data_type` to proportions (`"PR"`) and set axis ranges to zoom in on the meaningful range.

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",
  xrange = c(0, 6500),
  yrange = c(0, 0.05)
)
```

::: callout-note
-   `data_type = "PR"` changes from the default `"SR"` to proportions.
-   `xrange` and `yrange` set the range of x-axis and y-axis.
:::

### 4.3 Makeover 2

I further clean up the plot by removing default outlier labels and adding proper titles.

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",
  xrange = c(0, 6500),
  yrange = c(0, 0.05),
  label = NA,
  title = "Cumulative COVID-19 Fatality Rate by Cumulative Total Number of COVID-19 Positive Cases",
  x_label = "Cumulative COVID-19 Positive Cases",
  y_label = "Cumulative Fatality Rate"
)
```

::: callout-note
-   `label = NA` removes the default outlier labels.
-   `title`, `x_label` and `y_label` are used to add proper titles to the plot and axes.
:::

## 5. Funnel Plot with ggplot2

In this section, I build the funnel plot step-by-step from scratch using ggplot2 for full customisation control.

### 5.1 Computing the Basic Derived Fields

First, I derive the cumulative death rate and its standard error.

```{r}
df <- covid19 %>%
  mutate(rate = Death / Positive) %>%
  mutate(rate.se = sqrt((rate * (1 - rate)) / (Positive))) %>%
  filter(rate > 0)
```

Then I compute the weighted mean rate:

```{r}
fit.mean <- weighted.mean(df$rate, 1 / df$rate.se^2)
```

### 5.2 Calculate Lower and Upper Limits for 95% and 99.9% CI

```{r}
number.seq <- seq(1, max(df$Positive), 1)
number.ll95 <- fit.mean - 1.96 * sqrt((fit.mean * (1 - fit.mean)) / (number.seq))
number.ul95 <- fit.mean + 1.96 * sqrt((fit.mean * (1 - fit.mean)) / (number.seq))
number.ll999 <- fit.mean - 3.29 * sqrt((fit.mean * (1 - fit.mean)) / (number.seq))
number.ul999 <- fit.mean + 3.29 * sqrt((fit.mean * (1 - fit.mean)) / (number.seq))
dfCI <- data.frame(number.ll95, number.ul95, number.ll999,
                   number.ul999, number.seq, fit.mean)
```

### 5.3 Plotting a Static Funnel Plot

```{r}
p <- ggplot(df, aes(x = Positive, y = rate)) +
  geom_point(aes(label = `Sub-district`),
             alpha = 0.4) +
  geom_line(data = dfCI,
            aes(x = number.seq,
                y = number.ll95),
            linewidth = 0.4,
            colour = "grey40",
            linetype = "dashed") +
  geom_line(data = dfCI,
            aes(x = number.seq,
                y = number.ul95),
            linewidth = 0.4,
            colour = "grey40",
            linetype = "dashed") +
  geom_line(data = dfCI,
            aes(x = number.seq,
                y = number.ll999),
            linewidth = 0.4,
            colour = "grey40") +
  geom_line(data = dfCI,
            aes(x = number.seq,
                y = number.ul999),
            linewidth = 0.4,
            colour = "grey40") +
  geom_hline(data = dfCI,
             aes(yintercept = fit.mean),
             linewidth = 0.4,
             colour = "grey40") +
  coord_cartesian(ylim = c(0, 0.05)) +
  annotate("text", x = 1, y = -0.13, label = "95%", size = 3, colour = "grey40") +
  annotate("text", x = 4.5, y = -0.18, label = "99%", size = 3, colour = "grey40") +
  ggtitle("Cumulative Fatality Rate by Cumulative Number of COVID-19 Cases") +
  xlab("Cumulative Number of COVID-19 Cases") +
  ylab("Cumulative Fatality Rate") +
  theme_light() +
  theme(plot.title = element_text(size = 12),
        legend.position = c(0.91, 0.85),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 7),
        legend.background = element_rect(colour = "grey60", linetype = "dotted"),
        legend.key.height = unit(0.3, "cm"))
p
```

### 5.4 Interactive Funnel Plot: plotly + ggplot2

I convert the static ggplot2 funnel plot into an interactive one using `ggplotly()`.

```{r}
fp_ggplotly <- ggplotly(p,
                        tooltip = c("label", "x", "y"))
fp_ggplotly
```

## 6. References

-   [funnelPlotR](https://nhs-r-community.github.io/FunnelPlotR/index.html) package.
-   [Funnel Plots for Indirectly-standardised ratios](https://nhs-r-community.github.io/FunnelPlotR/articles/funnel_plots.html).
-   [Changing funnel plot options](https://nhs-r-community.github.io/FunnelPlotR/articles/changing_funnel_plot_options.html)
-   [ggplot2](https://ggplot2.tidyverse.org/) package.
