---
title: "Creating Ternary Plot with R"
date: "2026-05-10"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

------------------------------------------------------------------------

## Overview

Ternary plots are a way of displaying the distribution and variability of three-part compositional data. For example, the proportion of aged, economy active and young population, or sand, silt, and clay in soil. The display is a triangle with sides scaled from 0 to 1. Each side represents one of the three components. A point is plotted so that a line drawn perpendicular from the point to each leg of the triangle intersects at the component values of the point.

In this hands-on exercise, I will build ternary plots programmatically using R to visualise and analyse the population structure of Singapore.

My hands-on exercise consists of four steps:

-   I will install and launch **tidyverse** and **ggtern** packages.
-   I will derive three new measures using the *mutate()* function of the **dplyr** package.
-   I will build a static ternary plot using the *ggtern()* function of the **ggtern** package.
-   I will build an interactive ternary plot using the *plot_ly()* function of the **Plotly R** package.

## Installing and Launching R Packages

For this exercise, I will use two main R packages:

-   **ggtern**, a ggplot extension specially designed to plot ternary diagrams. The package will be used to plot static ternary plots.

-   [**Plotly R**](https://plot.ly/r/), an R package for creating interactive web-based graphs via plotly’s JavaScript graphing library, plotly.js . The **plotly R** libary contains the *ggplotly* function, which will convert **ggplot2** figures into a Plotly object.

I will also ensure that selected **tidyverse** family packages — namely **readr**, **dplyr**, and **tidyr** — are installed and loaded.

The code chunk below will accomplish this task.

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

## Data Preparation

### The Data

For this hands-on exercise, I will use the Singapore Residents by Planning AreaSubzone, Age Group, Sex and Type of Dwelling, June 2000-2018 data. The dataset has been downloaded and is stored in the Hands-on W5 sub-folder of my hands-on exercise folder. It is called *respopagsex2000to2018_tidy.csv* and is in CSV file format.

### Importing Data

To import *respopagsex2000to2018_tidy.csv* into R, I will use the ***read_csv()*** function of the **readr** package.

```{r}
#| eval: true
# Reading the data into R environment
pop_data <- read_csv("Hands-on Exercises/Hands-on W5/respopagsex2000to2018_tidy.csv")
```

### Preparing the Data

Next, I will use the ***mutate()*** function of the **dplyr** package to derive three new measures: `YOUNG`, `ACTIVE`, and `OLD`.

```{r}
#| eval: true
# Deriving the young, economy active and old measures
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year)) %>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8])) %>%
  mutate(ACTIVE = rowSums(.[9:16])) %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018) %>%
  filter(TOTAL > 0)
```

## Plotting Ternary Diagram with R

### Plotting a Static Ternary Diagram

I will use the ***ggtern()*** function of the **ggtern** package to create a simple ternary plot.

```{r}
#| eval: true
# Building the static ternary plot
ggtern(data = agpop_mutated, aes(x = YOUNG, y = ACTIVE, z = OLD)) +
  geom_point()
```

Next, I will enhance the plot by adding a title and applying the `theme_rgbw()` theme for a cleaner look.

```{r}
#| eval: true
# Building the static ternary plot with title and theme
ggtern(data = agpop_mutated, aes(x = YOUNG, y = ACTIVE, z = OLD)) +
  geom_point() +
  labs(title = "Population structure, 2018") +
  theme_rgbw()
```

### Plotting an Interactive Ternary Diagram

The code below creates an interactive ternary plot using the ***plot_ly()*** function of **Plotly R**. I first define two reusable helper functions — one for annotations and one for axis formatting — before initiating the visualisation.

```{r}
#| eval: true
# Reusable function for creating annotation object
label <- function(txt) {
  list(
    text = txt,
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper",
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# Reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"),
  baxis = axis("Active"),
  caxis = axis("Old")
)

# Initiating a plotly visualization
plot_ly(
  agpop_mutated,
  a = ~YOUNG,
  b = ~ACTIVE,
  c = ~OLD,
  color = I("black"),
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"),
    ternary = ternaryAxes
  )
```
