---
title: "Treemap Visualisation with R"
date: "2026-05-10"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## 1. Overview

In this hands-on exercise, I will gain hands-on experience designing treemaps using appropriate R packages. The exercise consists of three main sections. First, I will manipulate transaction data into a treemap structure using selected functions from the **dplyr** package. Then, I will plot static treemaps using the **treemap** package. In the third section, I will design interactive treemaps using the **d3treeR** package.

## 2. Installing and Launching R Packages

I will first check and load the **treemap**, **treemapify**, and **tidyverse** packages.

```{r}
pacman::p_load(treemap, treemapify, tidyverse)
```

## 3. Data Wrangling

In this exercise, I will use the *REALIS2018.csv* data, which provides information on private property transaction records in 2018. The dataset is extracted from the REALIS portal of the Urban Redevelopment Authority (URA).

### 3.1 Importing the Data Set

I will use *read_csv()* of **readr** to import realis2018.csv into R and parse it into a tibble data frame.

```{r}
realis2018 <- read_csv("Hands-on Exercises/Hands-on W5/realis2018.csv")
```

### 3.2 Data Wrangling and Manipulation

The *realis2018* data frame is in transaction record form — highly disaggregated and not appropriate for plotting a treemap. I will perform the following steps:

-   Group transaction records by *Project Name*, *Planning Region*, *Planning Area*, *Property Type*, and *Type of Sale*.
-   Compute *Total Unit Sold*, *Total Area*, *Median Unit Price*, and *Median Transacted Price* by applying appropriate summary statistics.

I will use two key **dplyr** verbs: *group_by()* and *summarise()*.

*group_by()* breaks down a data frame into specified groups of rows, after which verbs are automatically applied "by group". *summarise()* computes the summary for each group.

### 3.3 Grouped Summaries without the Pipe

The code chunk below shows a two-line approach to perform the steps.

```{r}
realis2018_grouped <- group_by(realis2018, `Project Name`,
                               `Planning Region`, `Planning Area`,
                               `Property Type`, `Type of Sale`)
realis2018_summarised <- summarise(realis2018_grouped,
                          `Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE),
                          `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
                          `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
                          `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

::: callout-note 
Aggregation functions such as *sum()* and *median()* follow the usual rule of missing values — if there's any missing value in the input, the output will be missing. The argument `na.rm = TRUE` removes missing values prior to computation.
::: 

### 3.4 Grouped Summaries with the Pipe

A more efficient way to tackle the same process is using the pipe `%>%`:

```{r}
realis2018_summarised <- realis2018 %>%
  group_by(`Project Name`, `Planning Region`,
           `Planning Area`, `Property Type`,
           `Type of Sale`) %>%
  summarise(`Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE),
            `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
            `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
            `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

## 4. Designing Treemap with treemap Package

The **treemap** package is specially designed to offer great flexibility in drawing treemaps. The core function *treemap()* offers at least 43 arguments. In this section, I will explore the major arguments for designing elegant and truthful treemaps.

### 4.1 Designing a Static Treemap

In this section, I will use *treemap()* to plot a treemap showing the distribution of median unit prices and total unit sold of resale condominiums by geographic hierarchy in 2018.

First, I will filter records belonging to resale condominium property type:

```{r}
realis2018_selected <- realis2018_summarised %>%
  filter(`Property Type` == "Condominium", `Type of Sale` == "Resale")
```

### 4.2 Using the Basic Arguments

The code chunk below designs a treemap using the three core arguments: *index*, *vSize*, and *vColor*.

```{r}
treemap(realis2018_selected,
        index = c("Planning Region", "Planning Area", "Project Name"),
        vSize = "Total Unit Sold",
        vColor = "Median Unit Price ($ psm)",
        title = "Resale Condominium by Planning Region and Area, 2018",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

Things to note:

-   `index`: the index vector must consist of at least two column names, otherwise no hierarchy treemap will be plotted. The first name is the highest aggregation level.
-   `vSize`: the column must not contain negative values, as its values map to the sizes of the rectangles.

::: callout-warning
The treemap above was wrongly coloured. For a correctly designed treemap, the colours of the rectangles should be in different intensities showing median unit prices. Without defining `type`, *treemap()* assumes `type = index`, meaning it colours by the hierarchy of planning areas.
:::

### 4.3 Working with vColor and type Arguments

I will now define the `type` argument as `"value"` for correct colouring:

```{r}
treemap(realis2018_selected,
        index = c("Planning Region", "Planning Area", "Project Name"),
        vSize = "Total Unit Sold",
        vColor = "Median Unit Price ($ psm)",
        type = "value",
        title = "Resale Condominium by Planning Region and Area, 2018",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

The rectangles are now coloured with different intensities of green, reflecting their respective median unit prices. The legend shows values binned into ten equal intervals of 5,000.

### 4.4 Colours in treemap Package

Two arguments determine the mapping to colour palettes: *mapping* and *palette*. The `"value"` treemap treats the palette as diverging (e.g. `"RdYlBu"`), mapping 0 to the middle colour, while `"manual"` maps min(values) to the left-end colour and max(values) to the right-end colour.

### 4.5 The "value" Type Treemap

```{r}
treemap(realis2018_selected,
        index = c("Planning Region", "Planning Area", "Project Name"),
        vSize = "Total Unit Sold",
        vColor = "Median Unit Price ($ psm)",
        type = "value",
        palette = "RdYlBu",
        title = "Resale Condominium by Planning Region and Area, 2018",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

Note that although the `RdYlBu` palette is used, there are no red rectangles — because all median unit prices are positive.

### 4.6 The "manual" Type Treemap

The `"manual"` type maps the value range linearly to the colour palette:

```{r}
treemap(realis2018_selected,
        index = c("Planning Region", "Planning Area", "Project Name"),
        vSize = "Total Unit Sold",
        vColor = "Median Unit Price ($ psm)",
        type = "manual",
        palette = "RdYlBu",
        title = "Resale Condominium by Planning Region and Area, 2018",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

The colour scheme is confusing here because a diverging palette like `RdYlBu` is not wise when all values are positive. I will use a single colour palette instead:

```{r}
treemap(realis2018_selected,
        index = c("Planning Region", "Planning Area", "Project Name"),
        vSize = "Total Unit Sold",
        vColor = "Median Unit Price ($ psm)",
        type = "manual",
        palette = "Blues",
        title = "Resale Condominium by Planning Region and Area, 2018",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### 4.7 Treemap Layout

*treemap()* supports two popular treemap layouts: `"squarified"` and `"pivotSize"`. The default is `"pivotSize"`. The squarified algorithm (Bruls et al., 2000) produces good aspect ratios but ignores sorting order, while the pivot-by-size algorithm (Bederson et al., 2002) takes sorting order into account.

### 4.8 Working with algorithm Argument

The code chunk below plots a squarified treemap:

```{r}
treemap(realis2018_selected,
        index = c("Planning Region", "Planning Area", "Project Name"),
        vSize = "Total Unit Sold",
        vColor = "Median Unit Price ($ psm)",
        type = "manual",
        palette = "Blues",
        algorithm = "squarified",
        title = "Resale Condominium by Planning Region and Area, 2018",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### 4.9 Using sortID

When `"pivotSize"` algorithm is used, the `sortID` argument determines the order in which rectangles are placed from top-left to bottom-right.

```{r}
treemap(realis2018_selected,
        index = c("Planning Region", "Planning Area", "Project Name"),
        vSize = "Total Unit Sold",
        vColor = "Median Unit Price ($ psm)",
        type = "manual",
        palette = "Blues",
        algorithm = "pivotSize",
        sortID = "Median Transacted Price",
        title = "Resale Condominium by Planning Region and Area, 2018",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

## 5. Designing Treemap using treemapify Package

**treemapify** is specially developed to draw treemaps in **ggplot2**. Before getting started, I recommend reading the [Introduction to treemapify](https://cran.r-project.org/web/packages/treemapify/vignettes/introduction-to-treemapify.html) and its [user guide](https://cran.r-project.org/web/packages/treemapify/treemapify.pdf).

### 5.1 Designing a Basic Treemap

```{r}
ggplot(data = realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`),
       layout = "scol",
       start = "bottomleft") +
  geom_treemap() +
  scale_fill_gradient(low = "light blue", high = "blue")
```

### 5.2 Defining Hierarchy

**Group by Planning Region:**

```{r}
ggplot(data = realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`),
       start = "topleft") +
  geom_treemap()
```

**Group by Planning Area:**

```{r}
ggplot(data = realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) +
  geom_treemap()
```

**Adding boundary lines:**

```{r}
ggplot(data = realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) +
  geom_treemap() +
  geom_treemap_subgroup2_border(colour = "gray40",
                                size = 2) +
  geom_treemap_subgroup_border(colour = "gray20")
```

## 6. Designing Interactive Treemap using d3treeR

### 6.1 Installing d3treeR Package

*d3treeR* is not available on CRAN. I will install it from GitHub as follows:

```{r}
#| eval: false
# Step 1: Install devtools if not already installed
install.packages("devtools")

# Step 2: Load devtools and install d3treeR from GitHub
library(devtools)
install_github("timelyportfolio/d3treeR")
```

```{r}
library(d3treeR)
```

### 6.2 Designing an Interactive Treemap

The code below performs two steps:

**Step 1:** I use *treemap()* to build a treemap using selected variables, saving the result as ***tm***.

```{r}
tm <- treemap(realis2018_summarised,
        index = c("Planning Region", "Planning Area"),
        vSize = "Total Unit Sold",
        vColor = "Median Unit Price ($ psm)",
        type = "value",
        title = "Private Residential Property Sold, 2018",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

**Step 2:** I use *d3tree()* to build an interactive treemap from the *tm* object.

```{r}
d3tree(tm, rootname = "Singapore")
```
