---
title: "Visual Multivariate Analysis with Parallel Coordinates Plot"
date: "2026-05-10"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## 1. Overview

Parallel coordinates plot is a data visualisation specially designed for visualising and analysing multivariate, numerical data. It is ideal for comparing multiple variables together and seeing the relationships between them — for example, the variables contributing to the Happiness Index. Parallel coordinates was invented by [Alfred Inselberg](http://www.math.tau.ac.il/~aiisreal/) in the 1970s as a way to visualise high-dimensional data. As pointed out by [Stephen Few](https://www.perceptualedge.com/articles/b-eye/parallel_coordinates.pdf) (2006), this technique's strength lies in its ability to bring meaningful multivariate patterns and comparisons to light when used interactively for analysis — for example, to characterise clusters detected during customer segmentation.

By the end of this hands-on exercise, I will have gained hands-on experience in:

- Plotting static parallel coordinates plots using [`ggparcoord()`](https://ggobi.github.io/ggally/reference/ggparcoord.html) of the [**GGally**](https://ggobi.github.io/ggally/) package, and
- Plotting interactive parallel coordinates plots using the [**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/) package.

## 2. Installing and Launching R Packages

For this exercise, I will use the **GGally**, **parallelPlot**, and **tidyverse** packages.

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```

## 3. Data Preparation

In this hands-on exercise, I will use the [World Happiness 2018](http://worldhappiness.report/ed/2018/) data. The original data set is in Microsoft Excel format and has been extracted and saved as **WHData-2018.csv**.

I will use `read_csv()` of the **readr** package to import *WHData-2018.csv* into R and save it as a tibble data frame called *wh*.

```{r}
wh <- read_csv("Hands-on Exercises/Hands-on W5/WHData-2018.csv")
```

## 4. Plotting Static Parallel Coordinates Plot

In this section, I will plot static parallel coordinates plots using [`ggparcoord()`](https://ggobi.github.io/ggally/reference/ggparcoord.html) of the **GGally** package.

### 4.1 Plotting a Simple Parallel Coordinates

The code chunk below shows the typical syntax to plot a basic static parallel coordinates plot using `ggparcoord()`.

```{r}
ggparcoord(data = wh,
           columns = c(7:12))
```

Only two arguments are used: `data` maps the data object (*wh*), and `columns` selects the columns for the plot.

### 4.2 Plotting a Parallel Coordinates with Boxplot

The basic parallel coordinates failed to reveal meaningful insights into the World Happiness measures. I will now improve the plot using a collection of arguments provided by `ggparcoord()`.

```{r}
ggparcoord(data = wh,
           columns = c(7:12),
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE,
           title = "Parallel Coordinates Plot of World Happiness Variables")
```

Things to note:

- `groupColumn` groups the parallel lines by a single variable (Region) and colours them by region name.
- `scale = "uniminmax"` univariately scales each variable so the minimum is 0 and the maximum is 1.
- `alphaLines = 0.2` reduces the intensity of the line colour. Permissible range is 0 to 1.
- `boxplot = TRUE` turns on the boxplot overlay.
- `title` provides the plot a title.

### 4.3 Parallel Coordinates with Facet

Since `ggparcoord()` is built on **ggplot2**, I can combine it with ggplot2 functions. Below I use `facet_wrap()` to plot 10 small multiple parallel coordinates plots — one per geographical region.

```{r}
ggparcoord(data = wh,
           columns = c(7:12),
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE,
           title = "Multiple Parallel Coordinates Plots of World Happiness Variables by Region") +
  facet_wrap(~ Region)
```

One aesthetic issue is that some variable names overlap on the x-axis.

### 4.4 Rotating x-axis Text Label

To make the x-axis text labels easier to read, I will rotate them by 30 degrees using `theme()`.

```{r}
ggparcoord(data = wh,
           columns = c(7:12),
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE,
           title = "Multiple Parallel Coordinates Plots of World Happiness Variables by Region") +
  facet_wrap(~ Region) +
  theme(axis.text.x = element_text(angle = 30))
```

`axis.text.x` is passed to `theme()` with `element_text(angle = 30)` to rotate x-axis text by 30 degrees.

### 4.5 Adjusting the Rotated x-axis Text Label

Rotating to 30 degrees may cause labels to overlap with the plot. I can avoid this by adjusting the text location using `hjust` within `element_text()`.

```{r}
ggparcoord(data = wh,
           columns = c(7:12),
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE,
           title = "Multiple Parallel Coordinates Plots of World Happiness Variables by Region") +
  facet_wrap(~ Region) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## 5. Plotting Interactive Parallel Coordinates Plot: parallelPlot Methods

[**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/) is an R package specially designed to plot parallel coordinates plots using the `htmlwidgets` package and [d3.js](https://d3js.org/). In this section, I will use functions provided in parallelPlot to build interactive parallel coordinates plots.

### 5.1 The Basic Plot

The code chunk below plots an interactive parallel coordinates plot using `parallelPlot()`.

```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250)
```

Some axis labels may be too long. I will address this in the next step.

### 5.2 Rotate Axis Label

I use the `rotateTitle` argument to avoid overlapping axis labels.

```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```

One useful interactive feature of parallelPlot is that clicking on a variable of interest (e.g. Happiness score) changes the colour scheme from a monotonous blue to blues with varying intensity.

### 5.3 Changing the Colour Scheme

I can change the default blue colour scheme using the `continuousCS` argument.

```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)
```

### 5.4 Parallel Coordinates Plot with Histogram

Using the `histoVisibility` argument, I can plot a histogram along the axis of each variable.

```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```

## 6. References

- [*ggparcoord()*](http://ggobi.github.io/ggally/reference/ggparcoord.html) of [**GGally**](http://ggobi.github.io/ggally/index.html) package
- [**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/parallelPlot.pdf)

