---
title: "Information Dashboard Design with R"
date: "2026-06-21"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: auto
---

```{=html}
<style>
td[headers="Monthly Sales"] svg g[clip-path] > rect {
  fill: none;
}

td[headers="Monthly Sales"] svg polyline {
  fill: none;
  stroke: #000;
}

td[headers="Actual"] svg line {
  stroke: #000;
}

td[headers="Product"],
th#Monthly-Sales {
  white-space: nowrap;
}
</style>
```

------------------------------------------------------------------------

## 1. Learning Outcome

By the end of this hands-on exercise, I will be able to:

-   create bullet charts using **ggplot2**,
-   create sparklines using **ggplot2**,
-   build a static information dashboard by using **gt** and **gtExtras**, and
-   build an interactive information dashboard by using **reactable** and **reactablefmtr**.

## 2. Getting Started

### 2.1 Installing and Loading R Packages

The R packages required for this exercise are: **lubridate**, **ggthemes**, **reactable**, **reactablefmtr**, **gt**, **gtExtras**, and **tidyverse**. The code chunk below uses `pacman::p_load()` to check, install, and load all the necessary packages.

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
               reactablefmtr, gt, gtExtras, tidyverse, dataui)
theme_set(theme_minimal())
```

## 3. Importing Microsoft Access Database

### 3.1 The Data Set

The data set used in this hands-on exercise is the *Coffee Chain* Microsoft Access database (*Coffee Chain.mdb*). It contains transactional sales data for a coffee chain, including product names, sales figures, budget targets, and dates.

### 3.2 Importing the Database into R

::: callout-note
The code below uses the **RODBC** package to connect to the Microsoft Access database. This approach only works on **Windows** with a compatible 32-bit ODBC driver. On macOS, this step is not available — I will load the pre-exported *CoffeeChain.rds* file directly instead (see the next code chunk).
:::

```{r}
#| eval: false
library(RODBC)
con <- odbcConnectAccess2007('Hands-on Exercises/Hands-on W10/Coffee Chain.mdb')
coffeechain <- sqlFetch(con, 'CoffeeChain Query')
write_rds(coffeechain, "Hands-on Exercises/Hands-on W10/CoffeeChain.rds")
odbcClose(con)
```

Since I am working on macOS, I will load the data directly from the pre-exported RDS file.

```{r}
coffeechain <- read_rds("Hands-on Exercises/Hands-on W10/CoffeeChain.rds")
```

### 3.3 Data Preparation

The code chunk below aggregates the data at the product level, deriving the total budget sales (`target`) and actual sales (`current`) for each product.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```

### 3.4 Bullet Chart in ggplot2

A bullet chart is a variation of a bar chart designed to replace dashboard gauges and meters. It compares an actual value against a target and shows qualitative performance ranges (e.g. poor, satisfactory, good) in the background.

```{r}
ggplot(product, aes(Product, current)) +
  geom_col(aes(Product, max(target) * 1.01),
           fill = "grey85", width = 0.85) +
  geom_col(aes(Product, target * 0.75),
           fill = "grey60", width = 0.85) +
  geom_col(aes(Product, target * 0.5),
           fill = "grey50", width = 0.85) +
  geom_col(aes(Product, current),
           width = 0.35,
           fill = "black") +
  geom_errorbar(aes(y = target,
                    x = Product,
                    ymin = target,
                    ymax = target),
                width = 0.4,
                colour = "red",
                size = 1) +
  coord_flip()
```

::: callout-note
## Things to learn from the code chunk

-   Three background bars of decreasing shade (grey85 → grey60 → grey50) represent qualitative performance bands — here set at 100%, 75%, and 50% of the target.
-   The narrow black bar represents actual current sales.
-   `geom_errorbar()` with `ymin = ymax = target` draws a single horizontal line marking the target value.
-   `coord_flip()` rotates the chart so products appear on the y-axis, making it easier to read long product names.
:::

## 4. Plotting Sparklines Using ggplot2

A sparkline is a small, condensed line chart without axes or labels that shows a trend at a glance. They are often placed alongside data tables to give context to summary statistics.

### 4.1 Preparing the Data

The code chunk below filters the data to 2013, derives a `Month` field, and aggregates sales by month and product.

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

Next, I compute the minimum and maximum sales months, the end-of-period values, and the interquartile range for each product — all of which will be used to annotate the sparklines.

```{r}
mins <- group_by(sales_report, Product) %>%
  slice(which.min(Sales))

maxs <- group_by(sales_report, Product) %>%
  slice(which.max(Sales))

ends <- group_by(sales_report, Product) %>%
  filter(Month == max(Month))

quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 0.25),
            quart2 = quantile(Sales, 0.75)) %>%
  right_join(sales_report)
```

### 4.2 Sparklines in ggplot2

The code chunk below uses `facet_grid()` to produce one sparkline per product, with an IQR ribbon, annotated min/max points, and end-of-period labels.

```{r}
ggplot(sales_report, aes(x = Month, y = Sales)) +
  facet_grid(Product ~ ., scales = "free_y") +
  geom_ribbon(data = quarts,
              aes(ymin = quart1, max = quart2),
              fill = "grey90") +
  geom_line(size = 0.3) +
  geom_point(data = mins, col = "red") +
  geom_point(data = maxs, col = "blue") +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) +
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        strip.text = element_blank())
```

::: callout-note
## Things to learn from the code chunk

-   `facet_grid(Product ~ ., scales = "free_y")` creates one panel per product with independent y-axis scales, so small-sales products are still readable.
-   The grey ribbon from `geom_ribbon()` visualises the interquartile range (Q1 to Q3) as a band of typical performance.
-   Red points mark monthly sales lows; blue points mark highs.
-   `expand_limits()` adds extra space on the right so the product name labels are not clipped.
-   `theme_tufte()` strips away all chart junk, keeping only the data elements.
:::

## 5. Static Information Dashboard: gt and gtExtras Methods

### 5.1 Plotting a Simple Bullet Chart with gt

The **gtExtras** package extends **gt** tables with chart functions. `gt_plt_bullet()` embeds a bullet chart directly into a table column.

```{r}
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current,
                target = target,
                width = 60,
                palette = c("lightblue", "black")) %>%
  gt_theme_538()
```

::: callout-note
-   `gt_plt_bullet()` takes the `current` column as the performance bar and `target` as the reference line.
-   `gt_theme_538()` applies a clean FiveThirtyEight-inspired table style.
:::

## 6. Sparklines: gtExtras Method

### 6.1 Plotting Coffeechain Sales Report

To use `gt_plt_sparkline()`, the monthly sales values must first be stored as a list column — one list of values per product row.

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate(Month = month(Date,
                       label = TRUE,
                       abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

The code chunk below converts the monthly sales into a list column.

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop")
```

The code chunk below plots the sparklines inside a gt table.

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop") %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

### 6.2 Adding Statistics

The code chunk below creates a summary statistics table showing the minimum, maximum, and average monthly sales per product.

```{r}
report %>%
  group_by(Product) %>%
  summarise("Min" = min(Sales, na.rm = TRUE),
            "Max" = max(Sales, na.rm = TRUE),
            "Average" = mean(Sales, na.rm = TRUE)) %>%
  gt() %>%
  fmt_number(columns = 4,
             decimals = 2)
```

### 6.3 Combining the Data Frames

I will now create two separate data frames — one with the sparkline list column and one with the summary statistics — and join them together.

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop")

sales <- report %>%
  group_by(Product) %>%
  summarise("Min" = min(Sales, na.rm = TRUE),
            "Max" = max(Sales, na.rm = TRUE),
            "Average" = mean(Sales, na.rm = TRUE))

sales_data <- left_join(sales, spark)
```

### 6.4 Plotting the Updated Data Table

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE) %>%
  cols_width(
    Product ~ pct(30),
    Min ~ pct(17),
    Max ~ pct(17),
    Average ~ pct(18),
    `Monthly Sales` ~ pct(18)
  ) %>%
  tab_options(table.width = pct(100))
```

### 6.5 Combining Bullet Chart and Sparklines

Finally, I will bring in the budget target data and combine the bullet chart and sparklines into a single dashboard-style table.

```{r}
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup()

sales_data <- sales_data %>%
  left_join(bullet)
```

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE) %>%
  gt_plt_bullet(column = Actual,
                target = Target,
                width = 28,
                palette = c("lightblue", "black")) %>%
  gt_theme_538() %>%
  cols_width(
    Product ~ pct(24),
    Min ~ pct(14),
    Max ~ pct(14),
    Average ~ pct(16),
    `Monthly Sales` ~ pct(18),
    Actual ~ pct(14)
  ) %>%
  tab_options(table.width = pct(100))
```

::: callout-note
This final static table combines three information layers in one compact view: summary statistics (Min, Max, Average), a sparkline showing the monthly sales trend, and a bullet chart comparing actual versus target sales — all the key elements of an information dashboard.
:::

## 7. Interactive Information Dashboard: reactable and reactablefmtr Methods

The **reactable** and **reactablefmtr** packages allow the creation of interactive HTML tables with embedded charts. The `react_sparkline()` function requires the **dataui** package, which must be installed from GitHub.

```{r}
remotes::install_github("timelyportfolio/dataui")
```

First, I will reshape the `report` data into the list-column format required by **reactablefmtr**.

```{r}
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))
```

### 7.1 Plotting Interactive Sparklines

```{r}
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

### 7.2 Changing the Page Size

By default, **reactable** shows 10 rows per page. Since there are 13 products, I will set `defaultPageSize = 13` to display all of them at once.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

### 7.3 Adding Points and Labels

The code chunk below highlights the minimum and maximum sales months with coloured points and adds first/last value labels.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
      )
    )
  )
)
```

### 7.4 Adding a Reference Line

The code chunk below adds a mean reference line to each sparkline, making it easy to see which months were above or below average.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
      )
    )
  )
)
```

### 7.5 Adding a Bandline

The code chunk below adds an interquartile range band (Q1 to Q3) in green, providing a visual reference for the typical range of monthly sales.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
      )
    )
  )
)
```

### 7.6 Changing from Sparkline to Sparkbar

Instead of a line chart, I can use `react_sparkbar()` to render each product's monthly trend as a bar chart.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean"
      )
    )
  )
)
```

::: callout-note
## Things to learn from the code chunk

-   `react_sparkbar()` replaces `react_sparkline()` to render bars instead of a line.
-   `highlight_bars()` colours the minimum month red and the maximum month blue, mirroring the point highlighting used in the sparkline version.
-   Combining `bandline` and `statline` gives readers both a typical-range band and a mean reference at a glance.
:::

## 8. Reference
