---
title: "Motor Vehicle Insurance Portfolio Visual Analytics Report"
subtitle: "ISSS608 Take-Home Exercise 1 - Technical Report"
date: "2026-05-10"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  message: false
  freeze: true
---

```{r}
#| label: global-theme
#| include: false

library(ggplot2)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(colour = "grey40", size = 11),
      plot.caption  = element_text(colour = "grey60", size = 9),
      plot.margin   = margin(12, 15, 12, 12)
    )
)
```

## 1. Overview

This technical report applies visual analytics techniques to a real-world **motor vehicle insurance portfolio** dataset from a Spanish non-life insurer (sourced from [Mendeley Data](https://data.mendeley.com/datasets/sw4jmdb2sm/1)). The dataset contains approximately 354,000 policy-year records spanning 2022-2024.

The two primary analytical objectives are:

-   **Univariate analysis**: Explore the distributions and characteristics of numerical and categorical variables across the portfolio.

-   **Bivariate analysis**: Investigate and statistically validate factors that influence **portfolio profitability** (loss ratio) and **volatility** (claim frequency and severity).

------------------------------------------------------------------------

## 2. Data Preparation

### 2.1 Approach

Before analysis, three preparatory steps were applied to the raw data:

-   **Type conversion**: Several variables are stored as numeric codes in the source file and were recoded as ordered or nominal factors with descriptive labels to support meaningful visualisation.

-   **Feature engineering**: 3 key actuarial performance metrics (loss ratio, claim frequency, claim severity) will be derived.

-   **Missing value assessment**

### 2.2. Loading Packages

```{r}
#| label: load-packages

pacman::p_load(
  tidyverse,    # data wrangling and ggplot2
  readxl,       # reading the data dictionary workbook
  ggridges,     # ridgeline density plots
  ggdist,       # half-eye and interval plots
  ggstatsplot,  # statistical annotations on plots
  ggExtra,      # marginal distributions on scatterplots
  GGally,       # parallel coordinates
  treemapify,   # treemap visualisation
  ggalluvial,   # alluvial flow visualisation
  patchwork,    # compositing multiple plots
  ggrepel,      # repelling text labels
  scales,       # axis formatting
  colorspace,   # colour palettes
  DT,           # interactive HTML tables
  broom,         # tidy statistical test outputs
  knitr         # table formatting
)
```

### 2.3 Data Import

The dataset uses a **semicolon (`;`) delimiter**, common in European data where commas serve as decimal separators. I use `read_delim()` with `delim = ";"` to handle this correctly.

```{r}
#| label: import-data
#| cache: true

motor_raw <- read_delim(
  "Take Home 1/A detailed dataset of motor insurance policies wit/Dataset of motor insurance portfolio.csv",
  delim = ";",
  show_col_types = FALSE
)

glimpse(motor_raw)
```

### 2.4 Data Dictionary

```{r}
#| label: data-dictionary

dict_path <- "Take Home 1/A detailed dataset of motor insurance policies wit/Descriptive of variables.xlsx"

if (file.exists(dict_path)) {
  dict_raw <- read_excel(dict_path, sheet = "Cartera")

  dict_clean <- dict_raw |>
    select(
      Variable = Variables,
      Description,
      Group
    ) |>
    filter(!is.na(Variable)) |>
    mutate(
      Group = replace_na(Group, "Continuation note")
    )
} else {
  dict_clean <- tibble(
    Variable = names(motor_raw),
    Description = NA_character_,
    Group = case_when(
      str_detect(Variable, "insured|year|policy|business|payment|bonus") ~
        "Policy and insured characteristics",
      str_detect(Variable, "driver|vehicle|fuel|seats|power|brand") ~
        "Driver and vehicle characteristics",
      str_detect(Variable, "municipality|circulation") ~
        "Geographic characteristics",
      str_detect(Variable, "premium") ~
        "Premium components",
      str_detect(Variable, "claims") ~
        "Claim count components",
      str_detect(Variable, "incurred") ~
        "Incurred loss components",
      str_detect(Variable, "exposure") ~
        "Exposure fields",
      TRUE ~ "Other fields"
    )
  )
}

dict_clean |>
  filter(Group != "Continuation note") |>
  count(Group, name = "Variables") |>
  knitr::kable(caption = "Variable Groups Identified from the Data Dictionary")
```

### 2.5 Initial Inspection

```{r}
#| label: inspect-missing

cat("Dimensions:", nrow(motor_raw), "rows ×", ncol(motor_raw), "columns\n\n")

missing_df <- motor_raw |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(),
               names_to  = "Variable",
               values_to = "Missing Count") |>
  filter(`Missing Count` > 0) |>
  mutate(`Missing (%)` = round(`Missing Count` / nrow(motor_raw) * 100, 2)) |>
  arrange(desc(`Missing Count`))

knitr::kable(missing_df, caption = "Variables with Missing Values")
```

Given the very low missingness rate, no row deletion or imputation was applied at this stage. Instead, observations with missing values were excluded only from analyses requiring those specific variables, preserving the maximum usable portfolio data for all other visual analytics tasks.

### 2.6 Reproducibility and Data Quality Checks

The source is a policy-year file, so a record should be uniquely identified by `insured_id` and `year`. The checks below verify row uniqueness, exposure bounds, missingness concentration, and whether the component premium and incurred fields reconcile closely with their totals.

```{r}
#| label: data-quality-checks

premium_components <- c(
  "liability_premium", "property_damage_premium", "theft_premium",
  "fire_premium", "glass_premium", "legal_protection_premium",
  "occupants_premium"
)

incurred_components <- c(
  "liability_incurred", "property_incurred", "theft_incurred",
  "fire_incurred", "glass_incurred", "legal_protection_incurred",
  "occupants_incurred"
)

dq_checks <- tibble(
  Check = c(
    "Policy-year duplicate records",
    "Rows with exposure outside [0, 1]",
    "Rows with non-positive total premium",
    "Rows with missing vehicle value",
    "Rows where premium components do not reconcile",
    "Rows where incurred components do not reconcile"
  ),
  Count = c(
    motor_raw |> count(insured_id, year) |> filter(n > 1) |> nrow(),
    motor_raw |> filter(total_exposure < 0 | total_exposure > 1) |> nrow(),
    motor_raw |> filter(total_premium <= 0) |> nrow(),
    sum(is.na(motor_raw$vehicle_value)),
    motor_raw |>
      mutate(component_premium = rowSums(across(all_of(premium_components)), na.rm = TRUE),
             diff = abs(total_premium - component_premium)) |>
      filter(diff > 0.01) |>
      nrow(),
    motor_raw |>
      mutate(component_incurred = rowSums(across(all_of(incurred_components)), na.rm = TRUE),
             diff = abs(total_incurred - component_incurred)) |>
      filter(diff > 0.01) |>
      nrow()
  )
) |>
  mutate(`Share of Rows` = percent(Count / nrow(motor_raw), 0.001))

knitr::kable(dq_checks, caption = "Analytical Sandbox Quality Checks")
```

### 2.7 Data Wrangling

Character columns are converted to factors with descriptive labels matching the data dictionary. `bonus_score` is declared as **ordered** (Good \> Neutral \> Bad) because it represents a ranked risk tier used in actuarial pricing.

```{r}
#| label: data-wrangling
#| cache: true

motor <- motor_raw |>
  mutate(
    year = factor(year),

    policy_type = factor(
      policy_type,
      levels = c("TP", "TPG", "CC", "COMP_E", "COMP_N")
    ),

    policy_status = factor(
      policy_status,
      levels = c("A", "C"),
      labels = c("Active", "Cancelled")
    ),

    business_type = factor(
      business_type,
      levels = c("NB", "P"),
      labels = c("New Business", "Portfolio")
    ),

    payment_frequency = factor(
      payment_frequency,
      levels = c("A", "S", "Q"),
      labels = c("Annual", "Semiannual", "Quarterly")
    ),

    bonus_score = factor(
      bonus_score,
      levels  = c("G", "N", "B"),
      labels  = c("Good", "Neutral", "Bad"),
      ordered = TRUE
    ),

    fuel_type = factor(
      fuel_type,
      levels = c("G", "D"),
      labels = c("Gasoline", "Diesel")
    ),

    municipality_type = factor(
      municipality_type,
      levels = c("I", "C", "IS"),
      labels = c("Inland", "Coastal", "Islands")
    ),

    circulation_area = factor(
      circulation_area,
      levels = c("U", "R"),
      labels = c("Urban", "Rural")
    ),

    vehicle_brand = str_to_title(vehicle_brand)
  )
```

### 2.8 Feature Engineering

Three core actuarial metrics and two derived grouping variables are created:

| Metric | Formula | Purpose |
|------------------------|------------------------|------------------------|
| **Loss Ratio** | `total_incurred / total_premium` | Profitability — values \> 1 indicate underwriting losses |
| **Claim Frequency** | `total_claims / total_exposure` | How often claims occur per unit of exposure |
| **Claim Severity** | `total_incurred / total_claims` | Average cost per claim (only for policies with ≥1 claim) |
| **Driving Experience** | `year − age_driving_licence` | Years since the driver obtained their licence |
| **Age Bands** | `cut()` on `driver_age` / `vehicle_age` | Discretised groups for segment-level analysis |

```{r}
#| label: feature-engineering
#| cache: true

motor <- motor |>
  mutate(
    loss_ratio = if_else(total_premium > 0,
                         total_incurred / total_premium,
                         NA_real_),

    claim_freq = if_else(total_exposure > 0,
                         total_claims / total_exposure,
                         NA_real_),

    claim_sev  = if_else(total_claims > 0,
                         total_incurred / total_claims,
                         NA_real_),

    has_claim  = total_claims > 0,

    driving_exp = as.integer(as.character(year)) - age_driving_licence,

    driver_age_band = cut(
      driver_age,
      breaks = c(17, 25, 35, 45, 55, 65, Inf),
      labels = c("18-25", "26-35", "36-45", "46-55", "56-65", "65+"),
      right  = TRUE
    ),

    vehicle_age_band = cut(
      vehicle_age,
      breaks = c(-Inf, 2, 5, 10, 15, Inf),
      labels = c("0-2", "3-5", "6-10", "11-15", "15+")
    )
  )
```

### 2.9 Portfolio Summary

```{r}
#| label: portfolio-summary

motor |>
  summarise(
    `Total Policy-Years`       = comma(n()),
    `Unique Insureds`          = comma(n_distinct(insured_id)),
    `Years Covered`            = paste(levels(year), collapse = ", "),
    `Active Policies`          = percent(mean(policy_status == "Active"), 0.1),
    `Policies with Claims`     = percent(mean(has_claim), 0.1),
    `Mean Loss Ratio`          = as.character(round(mean(loss_ratio, na.rm = TRUE), 4)),
    `Aggregate Loss Ratio`     = as.character(round(sum(total_incurred) / sum(total_premium), 4)),
    `Mean Total Premium (EUR)` = comma(round(mean(total_premium), 2))
  ) |>
  pivot_longer(everything(), names_to = "Metric", values_to = "Value") |>
  knitr::kable(caption = "Portfolio Summary Statistics")
```

### 2.10 Premium and Loss Component Sandbox

The portfolio has multiple coverage components. Before analysing overall loss ratio, the component view below checks which coverages contribute most to written premium and incurred losses. This is useful because volatility may be concentrated in one claim type even when aggregate portfolio profitability appears acceptable.

```{r}
#| label: component-sandbox

component_long <- motor |>
  transmute(
    year,
    Liability = liability_premium,
    Property = property_damage_premium,
    Theft = theft_premium,
    Fire = fire_premium,
    Glass = glass_premium,
    Legal = legal_protection_premium,
    Occupants = occupants_premium,
    Liability_incurred = liability_incurred,
    Property_incurred = property_incurred,
    Theft_incurred = theft_incurred,
    Fire_incurred = fire_incurred,
    Glass_incurred = glass_incurred,
    Legal_incurred = legal_protection_incurred,
    Occupants_incurred = occupants_incurred
  ) |>
  pivot_longer(
    -year,
    names_to = "component",
    values_to = "amount"
  ) |>
  mutate(
    measure = if_else(str_detect(component, "_incurred$"), "Incurred Loss", "Premium"),
    component = component |>
      str_remove("_incurred$") |>
      str_replace_all("_", " ") |>
      str_to_title()
  )

component_summary <- component_long |>
  group_by(year, measure, component) |>
  summarise(amount = sum(amount, na.rm = TRUE), .groups = "drop") |>
  group_by(year, measure) |>
  mutate(share = amount / sum(amount)) |>
  ungroup()
```

```{r}
#| label: component-composition
#| fig-cap: "Premium and incurred-loss composition by year"
#| fig-height: 7

component_summary |>
  ggplot(aes(x = year, y = share, fill = component)) +
  geom_col(width = 0.72, colour = "white", linewidth = 0.25) +
  facet_wrap(~measure) +
  scale_y_continuous(labels = percent) +
  scale_fill_brewer(palette = "Set3", name = "Coverage Component") +
  labs(
    title = "Portfolio Composition by Coverage Component",
    subtitle = "Comparing premium share with incurred-loss share identifies whether pricing and loss emergence are aligned",
    x = NULL,
    y = "Share of Annual Total"
  ) +
  theme(legend.position = "bottom")
```

**Interpretation**:

-   Liability and property dominate both premium and incurred losses, with the loss-side mix showing slightly more year-on-year variation than the premium side. This suggests aggregate loss ratios may mask component-level differences, motivating a closer examination of frequency and severity by coverage type.

------------------------------------------------------------------------

## 3. Univariate Visual Analytics

This section examines each variable individually to understand the portfolio's composition: the mix of coverage types, driver profiles, vehicle characteristics, and geographic spread, before moving to inter-variable comparisons.

### 3.1 Categorical Variables

#### 3.1.1 Policy and Insured Characteristics

Policy type is a nominal variable, so a sorted horizontal bar chart is used to compare category size directly while keeping the policy labels readable.

```{r}
#| label: cat-policy-type
#| fig-cap: "Distribution of policy types — CC and COMP_E are the dominant coverage structures"
#| fig-height: 5

motor |>
  count(policy_type) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = fct_reorder(policy_type, n), y = n, fill = policy_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(pct, 0.1)), hjust = -0.15, size = 4) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.18))) +
  scale_fill_brewer(palette = "Set2") +
  coord_flip() +
  labs(
    title    = "Distribution of Policy Types",
    subtitle = "CC and COMP_E (comprehensive with excess) account for the majority of the portfolio",
    x        = "Policy Type",
    y        = "Number of Policies",
    caption  = "TP: Third-party | TPG: TP + Glass | CC: TP + extras | COMP_E/N: Comprehensive"
  )
```

The remaining categorical policy attributes are shown as small-multiple bar charts so their distributions can be compared consistently without overloading a single axis.

```{r}
#| label: cat-policy-characteristics
#| fig-cap: "Policy-level categorical variables: status, business origin, payment frequency, and bonus history"
#| fig-height: 9

make_bar <- function(var, title, palette = "Blues") {
  motor |>
    count(!!sym(var)) |>
    mutate(pct = n / sum(n)) |>
    ggplot(aes(x = fct_reorder(!!sym(var), n), y = n, fill = !!sym(var))) +
    geom_col(show.legend = FALSE, alpha = 0.9) +
    geom_text(aes(label = percent(pct, 0.1)), hjust = -0.15, size = 3.5) +
    scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.22))) +
    scale_fill_brewer(palette = palette) +
    coord_flip() +
    labs(title = title, x = NULL, y = "Count") +
    theme(plot.title = element_text(face = "bold", size = 11))
}

p1 <- make_bar("policy_status",     "Policy Status",     "Blues")
p2 <- make_bar("business_type",     "Business Type",     "Greens")
p3 <- make_bar("payment_frequency", "Payment Frequency", "Oranges")
p4 <- make_bar("bonus_score",       "Bonus Score",       "Purples")

(p1 | p2) / (p3 | p4) +
  plot_annotation(
    title   = "Policy and Insured Characteristics",
    caption = "Source: Spanish Motor Insurance Portfolio (2022-2024)",
    theme   = theme(plot.title = element_text(face = "bold", size = 15))
  )
```

**Interpretation**:

-   The portfolio consists predominantly of **active policies** (89.2%) and annually-paying (81.3%) policies, reflecting a relatively clean operational structure. However, new business accounts for 69.3% of policy-years, which warrants caution when interpreting loss ratios — new business typically carries higher claim propensity due to the absence of renewal selection and policy seasoning effects, and profitability figures may not reflect long-run portfolio behaviour.

-   The bonus-score distribution is heavily concentrated in **"Good"** risks (96.4%), with Neutral and Bad segments accounting for just 2.2% and 1.3% respectively. While this may reflect genuine portfolio quality, it severely limits the variable's discriminatory power. Any comparison of loss experience across bonus-score categories should be treated with caution, as the exposure in non-Good segments is too thin for credible statistical inference.

#### 3.1.2 Vehicle and Geographic Characteristics

Fuel type, municipality type, and circulation area are compact categorical variables, making aligned bar charts suitable for quickly comparing their portfolio shares.

```{r}
#| label: cat-geo-fuel
#| fig-cap: "Fuel type, municipality type, and circulation area distributions"
#| fig-height: 5

p5 <- make_bar("fuel_type",         "Fuel Type",         "Reds")
p6 <- make_bar("municipality_type", "Municipality Type", "GnBu")
p7 <- make_bar("circulation_area",  "Circulation Area",  "Purples")

p5 | p6 | p7
```

#### 3.1.3 Vehicle Brand Composition - Treemap

A treemap is appropriate for vehicle brand because there are many categories; area makes the dominant exposure concentrations visible without requiring a long ranked bar chart.

```{r}
#| label: treemap-brand
#| fig-cap: "Top 25 vehicle brands — area and colour both encode policy count"
#| fig-height: 8

brand_counts <- motor |>
  count(vehicle_brand, sort = TRUE) |>
  slice_head(n = 25) |>
  mutate(pct = n / nrow(motor))

ggplot(
  brand_counts,
  aes(area  = n,
      fill  = n,
      label = paste0(vehicle_brand, "\n", comma(n)))
) +
  geom_treemap(colour = "white", linewidth = 0.5) +
  geom_treemap_text(
    colour = "white", place = "centre",
    size = 11, reflow = TRUE
  ) +
  scale_fill_gradient(
    low  = "#B3CDE3", high = "#08519C",
    name = "Policies", labels = comma
  ) +
  labs(
    title    = "Top 25 Vehicle Brands by Policy Count",
    subtitle = "Renault, Citroen, Peugeot, Volkswagen, Ford, Opel, and Seat form the main exposure base"
  )
```

**Interpretation**:

-   The brand exposure is broadly distributed, with no single manufacturer dominating the portfolio. Renault, Citroën, Peugeot, Volkswagen, Ford, Opel, and Seat collectively form the core of the book, each contributing between approximately 28,000 and 34,000 policies. Given their scale, loss experience within these brands will be the primary driver of overall portfolio results, brand-level underwriting attention should therefore focus on whether these high-volume makes exhibit adverse claim frequency or severity, rather than on the tail of smaller brands where credibility is limited.

-   Premium and near-premium brands: Mercedes, Audi, and BMW, are present but comparatively minor in policy count. While intuition suggests higher repair costs for these vehicles, any separate rating or loading applied to them should be grounded in observed loss evidence from this portfolio rather than assumed from general market benchmarks. Repair cost inflation and parts availability vary across manufacturers and model years, and without claim-level severity data by brand, cost-based assumptions risk being either overstated or misallocated.

### 3.2 Numerical Variables

#### 3.2.1 Premium Distribution

Most insurance premium distributions are right-skewed because a small proportion of high-risk or high-value policies attract substantially elevated premiums. The histogram shows the mass of annual premium levels, while the density overlay makes the skewness and possible multiple peaks easier to see.

```{r}
#| label: num-premium
#| fig-cap: "Total premium distribution is strongly right-skewed"
#| fig-height: 6

motor |>
  filter(total_premium > 0,
         total_premium < quantile(total_premium, 0.99)) |>
  ggplot(aes(x = total_premium)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins  = 80, fill = "#1565C0", alpha = 0.6
  ) +
  geom_density(colour = "#0D47A1", linewidth = 1.1) +
  scale_x_continuous(labels = dollar_format(prefix = "EUR ")) +
  labs(
    title    = "Distribution of Total Premium (capped at 99th percentile)",
    subtitle = "Right-skewed; most policies pay under EUR 800/year — COMP covers drive the upper tail",
    x        = "Total Annual Premium (EUR)",
    y        = "Density"
  )
```

**Interpretation**:

-   Most policyholders pay under EUR 300 annually, suggesting the portfolio is predominantly low-premium, high-volume in nature, which is a business model sensitive to loss frequency rather than severity. The right tail indicates a smaller segment of comprehensively covered, likely higher-value vehicles where individual claim severity matters more.

<!-- -->

-   The near-bimodal shape implies the portfolio is effectively serving two distinct customer segments with different coverage needs and risk profiles, yet they are being analysed and possibly priced as a single book. This has implications for profitability monitoring, aggregate loss ratios may mask divergent performance between the low-premium mass and the high-premium tail.

-   From a business perspective, growth strategy and retention efforts should distinguish between these segments, as the economics of acquiring and retaining a EUR 150 policy versus a EUR 700 policy are materially different in terms of expense ratio, claims handling cost, and margin contribution.

#### 3.2.2 Loss Ratio Distribution with Statistical Test

The loss ratio (incurred losses ÷ earned premium) is the primary profitability metric in non-life insurance. A value below 1.0 indicates underwriting profit; above 1.0 indicates a loss. I test whether the portfolio median loss ratio is significantly below the break-even threshold of 1.0 using a one-sample Wilcoxon signed-rank test, since the distribution is non-normal.

```{r}
#| label: histostats-lr
#| fig-cap: "Loss ratio distribution with embedded Wilcoxon test against break-even (LR = 1)"
#| fig-height: 7

motor_lr <- motor |>
  filter(!is.na(loss_ratio), loss_ratio >= 0,
         loss_ratio <= 5, total_premium > 0)

set.seed(42)
gghistostats(
  data         = motor_lr |> slice_sample(n = 50000),
  x            = loss_ratio,
  test.value   = 1,
  type         = "nonparametric",
  normal.curve = TRUE,
  title        = "Distribution of Policy-Level Loss Ratio",
  xlab         = "Loss Ratio (Incurred / Premium)",
  bar.fill     = "#EF9A9A"
)
```

**Interpretation**:

-   The distribution is heavily **zero-inflated**: the large mass at zero reflects the majority of policies that incur no claims, and exhibits a long right tail driven by a minority of policies with large individual losses. The Wilcoxon test confirms that the typical policy-year sits below break-even, but aggregate profitability should be read from the portfolio loss ratio itself: incurred losses are about 70% of premium, leaving the book profitable while still exposed to tail volatility.

-   From an insurance and reinsurance perspective, the zero mass and long right tail must be read together. The portfolio can be profitable on average while still exposing the insurer to retained earnings volatility if a small number of claim-bearing policy-years dominate incurred losses. The diagnostics below therefore separate attritional behaviour from large-loss concentration before moving into segment comparisons.

```{r}
#| label: reinsurance-tail-summary

claim_tail <- motor |>
  filter(total_incurred > 0) |>
  arrange(desc(total_incurred)) |>
  mutate(
    rank = row_number(),
    claim_policy_share = rank / n(),
    incurred_share = cumsum(total_incurred) / sum(total_incurred)
  )

tail_summary <- tibble(
  Metric = c(
    "Aggregate loss ratio",
    "Policy-years with at least one claim",
    "Exposure-adjusted claim frequency",
    "Portfolio average severity",
    "Top 1% of claim-bearing policy-years: incurred share",
    "Top 5% of claim-bearing policy-years: incurred share",
    "Top 10% of claim-bearing policy-years: incurred share",
    "99th percentile incurred loss, claim-bearing rows",
    "99.5th percentile incurred loss, claim-bearing rows",
    "99.9th percentile incurred loss, claim-bearing rows",
    "Maximum policy-year incurred loss"
  ),
  Value = c(
    percent(sum(motor$total_incurred) / sum(motor$total_premium), 0.1),
    percent(mean(motor$has_claim), 0.1),
    percent(sum(motor$total_claims) / sum(motor$total_exposure), 0.1),
    dollar(sum(motor$total_incurred) / sum(motor$total_claims), prefix = "EUR ", accuracy = 1),
    percent(sum(claim_tail$total_incurred[claim_tail$rank <= ceiling(nrow(claim_tail) * 0.01)]) / sum(claim_tail$total_incurred), 0.1),
    percent(sum(claim_tail$total_incurred[claim_tail$rank <= ceiling(nrow(claim_tail) * 0.05)]) / sum(claim_tail$total_incurred), 0.1),
    percent(sum(claim_tail$total_incurred[claim_tail$rank <= ceiling(nrow(claim_tail) * 0.10)]) / sum(claim_tail$total_incurred), 0.1),
    dollar(quantile(claim_tail$total_incurred, 0.99), prefix = "EUR ", accuracy = 1),
    dollar(quantile(claim_tail$total_incurred, 0.995), prefix = "EUR ", accuracy = 1),
    dollar(quantile(claim_tail$total_incurred, 0.999), prefix = "EUR ", accuracy = 1),
    dollar(max(claim_tail$total_incurred), prefix = "EUR ", accuracy = 1)
  )
)

knitr::kable(tail_summary, caption = "Tail Diagnostics for Portfolio Loss Volatility")
```

A Pareto-style cumulative curve is used here because concentration is easier to assess by ranking claim-bearing rows from largest to smallest rather than by reading individual losses.

```{r}
#| label: pareto-tail
#| fig-cap: "Pareto concentration of incurred losses among claim-bearing policy-years"
#| fig-height: 6

claim_tail |>
  ggplot(aes(x = claim_policy_share, y = incurred_share)) +
  geom_line(colour = "#B71C1C", linewidth = 1.3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = c(0.01, 0.05, 0.10), linetype = "dotted", colour = "grey40") +
  annotate(
    "text", x = 0.12, y = 0.55,
    label = "A small share of claim-bearing rows\ndrives a large share of incurred loss",
    hjust = 0, size = 4, colour = "#7F0000"
  ) +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Incurred Loss Concentration",
    subtitle = "The curve above the diagonal shows large-loss concentration; dotted lines mark the top 1%, 5%, and 10% claim-bearing rows",
    x = "Share of Claim-Bearing Policy-Years, Ranked by Incurred Loss",
    y = "Cumulative Share of Incurred Loss"
  )
```

**Critical reading**:

-   The portfolio is profitable in aggregate, but profitability is not evenly generated. The top claim-bearing policy-years carry a material share of incurred losses, and the maximum policy-year loss is many times the average severity. This means the underwriting discussion should not stop at "loss ratio below 100%"; it should ask which retained layers and segments create earnings volatility.

#### 3.2.3 Driver and Vehicle Characteristics - Ridgeline Plots

Comparing the shape of a distribution across multiple groups simultaneously is essential for detecting risk segmentation opportunities.

```{r}
#| label: ridgeline-age-by-year
#| fig-cap: "Driver age distribution across years — stable cohort composition"
#| fig-height: 6

motor |>
  filter(!is.na(driver_age)) |>
  ggplot(aes(x = driver_age, y = year, fill = after_stat(x))) +
  geom_density_ridges_gradient(
    scale = 1.5, rel_min_height = 0.01,
    colour = "white", linewidth = 0.4
  ) +
  scale_fill_viridis_c(option = "plasma", name = "Driver Age") +
  scale_x_continuous(breaks = seq(20, 90, 10)) +
  labs(
    title    = "Driver Age Distribution by Year",
    subtitle = "The insured driver cohort remains stable in composition across 2022-2024",
    x        = "Driver Age (years)",
    y        = "Year"
  ) +
  theme_ridges(font_size = 13) +
  theme(
    plot.title       = element_text(face = "bold"),
    legend.position  = "none"
  )
```

A paired ridgeline layout extends the same distribution-comparison logic to vehicle age and value, allowing coverage classes to be compared on both dimensions side by side.

```{r}
#| label: ridgeline-vehicle-chars
#| fig-cap: "Vehicle age (left) and vehicle value (right) distributions by policy type"
#| fig-height: 7

p_vage <- motor |>
  filter(!is.na(vehicle_age)) |>
  ggplot(aes(x = vehicle_age, y = policy_type, fill = policy_type)) +
  geom_density_ridges(alpha = 0.7, scale = 1.5, show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Vehicle Age by Policy Type",
    x     = "Vehicle Age (years)", y = NULL
  ) +
  theme_ridges(font_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 11))

p_vval <- motor |>
  filter(!is.na(vehicle_value), vehicle_value > 0,
         vehicle_value < quantile(vehicle_value, 0.99, na.rm = TRUE)) |>
  ggplot(aes(x = vehicle_value, y = policy_type, fill = policy_type)) +
  geom_density_ridges(alpha = 0.7, scale = 1.5, show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  scale_x_continuous(labels = dollar_format(prefix = "EUR ")) +
  labs(
    title = "Vehicle Value by Policy Type",
    x     = "Vehicle Value (EUR)", y = NULL
  ) +
  theme_ridges(font_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 11))

p_vage | p_vval
```

**Interpretation**:

-   Comprehensive policies (COMP_E and COMP_N) are predominantly associated with newer and higher-value vehicles, consistent with the rational economic behaviour of owners seeking to protect assets with meaningful residual value. This explains their elevated premium structure, but it also implies higher average claim severity, as newer vehicles carry greater parts and repair costs, and total loss settlements are more expensive. Reserving assumptions for this segment should reflect current vehicle values rather than portfolio-wide averages.

-   Basic TP and TPG policies, by contrast, cluster around older and lower-value vehicles, representing a structurally different risk pool in terms of both exposure and claim economics. While average severity per claim is lower, older vehicles may contribute to higher at-fault frequency through mechanical reliability issues, and their lower insured values reduce the deterrent effect of financial loss on driver behaviour. These two segments should be priced, reserved, and monitored separately, pooling them with comprehensive policies in aggregate analysis risks cross-subsidisation and masks divergent loss trends that would otherwise be actionable.

#### 3.2.4 Claim Severity Distribution by Policy Type

Understanding how claim costs are distributed matters for two core business decisions: how much to reserve for outstanding claims, and where to set reinsurance attachment points. The half-eye and boxplot combination is used because severity is skewed: it shows the full distribution shape while still retaining familiar median and interquartile-range summaries.

```{r}
#| label: raincloud-severity
#| fig-cap: "Claim severity distribution by policy type - right-skewed costs and uneven upper tails"
#| fig-height: 7

motor |>
  filter(!is.na(claim_sev), claim_sev > 0,
         claim_sev < quantile(claim_sev, 0.975, na.rm = TRUE)) |>
  ggplot(aes(x = claim_sev, y = policy_type,
             fill = policy_type, colour = policy_type)) +
  stat_halfeye(
    adjust = 0.5, width = 0.6,
    justification = -0.3, .width = 0, point_colour = NA
  ) +
  geom_boxplot(
    width      = 0.15, outlier.shape = NA,
    fill       = "white", colour = "black", linewidth = 0.5
  ) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  scale_colour_brewer(palette = "Set2", guide = "none") +
  scale_x_continuous(labels = dollar_format(prefix = "EUR ")) +
  labs(
    title    = "Claim Severity Distribution by Policy Type",
    subtitle = "Severity is right-skewed in every policy type; upper-tail behaviour differs more than the medians",
    x        = "Average Incurred Cost per Claim (EUR)",
    y        = "Policy Type"
  )
```

**Interpretation**:

-   Comprehensive policies (COMP_N and COMP_E) show wider interquartile ranges and heavier right tails, meaning claim costs are not only higher on average but more variable, directly increasing reserving uncertainty and reinsurance exposure. CC, as the largest policy type, shows similar dispersion, making its severity variability the most consequential for portfolio earnings.

-   TPG and TP are tighter and more predictable, though TP's isolated outliers beyond EUR 3,000 warrant monitoring given how far they deviate from the core distribution. The multimodal shapes across several policy types suggest claims are not homogeneous within each category, likely reflecting minor repairs and major structural losses being pooled together. A single average severity assumption per policy type would be inadequate; further segmentation by claim type is warranted for both pricing and reserving purposes.

------------------------------------------------------------------------

## 4. Bivariate Visual Analytics

All group comparisons use non-parametric Kruskal-Wallis and Dunn tests because the loss ratio and claim metrics are heavily right-skewed. Statistical test results are embedded directly in the plots so that visual patterns and inferential evidence are read together rather than separately.

> **Sampling note**: Plots using `ggbetweenstats()` draw a reproducible sample of 50,000 records (`set.seed(42)`) for rendering performance. Aggregate statistics quoted in the text are computed from the full 354,000-row dataset.

### 4.1 Exposure-Weighted Segment Scorecard

Policy-level averages can be misleading in insurance because exposure varies across cancelled and short-duration policies. The scorecard below aggregates premium, incurred loss, claims, and exposure by key underwriting cells. It ranks segments by *excess incurred loss* above the portfolio loss ratio benchmark.

```{r}
#| label: segment-scorecard

portfolio_lr <- sum(motor$total_incurred) / sum(motor$total_premium)

segment_scorecard <- motor |>
  filter(!is.na(driver_age_band), !is.na(bonus_score)) |>
  group_by(policy_type, bonus_score, driver_age_band) |>
  summarise(
    policies = n(),
    exposure = sum(total_exposure, na.rm = TRUE),
    premium = sum(total_premium, na.rm = TRUE),
    incurred = sum(total_incurred, na.rm = TRUE),
    claims = sum(total_claims, na.rm = TRUE),
    agg_lr = incurred / premium,
    claim_frequency = claims / exposure,
    avg_severity = if_else(claims > 0, incurred / claims, NA_real_),
    excess_incurred = incurred - portfolio_lr * premium,
    .groups = "drop"
  ) |>
  filter(policies >= 500, premium > 0, exposure > 0) |>
  arrange(desc(excess_incurred))

segment_scorecard |>
  slice_head(n = 12) |>
  transmute(
    `Policy Type` = policy_type,
    `Bonus Score` = bonus_score,
    `Driver Age Band` = driver_age_band,
    Policies = comma(policies),
    Exposure = comma(round(exposure, 1)),
    `Aggregate LR` = percent(agg_lr, 0.1),
    `Claim Frequency` = percent(claim_frequency, 0.1),
    `Avg Severity` = dollar(avg_severity, prefix = "EUR ", accuracy = 1),
    `Excess Incurred` = dollar(excess_incurred, prefix = "EUR ", accuracy = 1)
  ) |>
  knitr::kable(caption = "Top Adverse Underwriting Cells by Excess Incurred Loss")
```

### 4.2 Profitability Analysis - Loss Ratio

#### 4.2.1 Loss Ratio by Policy Type

A non-parametric comparison plot is used because policy-level loss ratios are zero-inflated and right-skewed; it combines distribution, central tendency, and test evidence in one view.

```{r}
#| label: biv-lr-ptype
#| fig-cap: "Loss ratio varies significantly across policy types — comprehensive covers are the most costly"
#| fig-height: 8

set.seed(42)
ggbetweenstats(
  data                 = motor_lr |> slice_sample(n = 50000),
  x                    = policy_type,
  y                    = loss_ratio,
  type                 = "nonparametric",
  p.adjust.method      = "holm",
  pairwise.comparisons = FALSE,
  centrality.plotting  = TRUE,
  point.args           = list(alpha = 0.01, size = 0.3),
  title                = "Loss Ratio by Policy Type",
  xlab                 = "Policy Type",
  ylab                 = "Loss Ratio (Incurred / Premium)"
)
```

**Interpretation**:

-   The Kruskal-Wallis test confirms that loss-ratio distributions differ by policy type, although the test does not imply every pair of policy types is materially different. The aggregate pattern is strongest for comprehensive covers: COMP_N has the highest aggregate loss ratio and COMP_E is also above the narrower TP/CC covers. Because policy-level medians are pulled to zero by no-claim records, the underwriting story should rely on aggregate LR, mean LR, claim frequency, and tail shape rather than median LR alone.

#### 4.2.2 Loss Ratio by Bonus Score

The bonus-malus system uses an insured's past claims history to classify their current risk tier. A key actuarial question is whether this classification actually predicts future loss outcomes. Pairwise Dunn tests between all three tiers directly test this assumption.

```{r}
#| label: biv-lr-bonus
#| fig-cap: "Each bonus tier differs significantly — the bonus-malus system discriminates risk effectively"
#| fig-height: 8

set.seed(42)
ggbetweenstats(
  data                 = motor_lr |> slice_sample(n = 50000),
  x                    = bonus_score,
  y                    = loss_ratio,
  type                 = "nonparametric",
  p.adjust.method      = "holm",
  pairwise.comparisons = TRUE,
  point.args           = list(alpha = 0.01, size = 0.3),
  title                = "Loss Ratio by Bonus Score",
  subtitle             = "Bonus tiers differ statistically, though aggregate LR is not perfectly ordered",
  xlab                 = "Bonus Score",
  ylab                 = "Loss Ratio"
)
```

**Interpretation**:

-   The bonus-malus groups differ statistically, but the practical signal is clearer for claim occurrence and frequency than for raw aggregate loss ratio. Good risks have the lowest aggregate LR and claim rate, while Neutral and Bad risks are both adverse; Bad risks show the highest claim frequency and severity even though Neutral produces the highest aggregate LR in this sample. This supports retaining bonus score as a rating variable, while avoiding an over-simple Good \< Neutral \< Bad loss-ratio narrative.

#### 4.2.3 Loss Ratio by Driver Age - Trend and Volatility

Driver age is a fundamental underwriting variable. Beyond the mean loss ratio by age band, the coefficient of variation (CV = standard deviation / mean) measures how *unpredictable* outcomes are within each group — a critical input for pricing uncertainty and capital allocation.

```{r}
#| label: biv-lr-driver-age
#| fig-cap: "Driver age affects claim frequency, aggregate loss ratio, and volatility in different ways"
#| fig-height: 8

lr_by_age <- motor_lr |>
  filter(!is.na(driver_age_band)) |>
  group_by(driver_age_band) |>
  summarise(
    n        = n(),
    agg_lr   = sum(total_incurred) / sum(total_premium),
    mean_lr  = mean(loss_ratio),
    se_lr    = sd(loss_ratio) / sqrt(n),
    cv_lr    = sd(loss_ratio) / mean(loss_ratio),
    .groups  = "drop"
  )

portfolio_lr <- sum(motor$total_incurred) / sum(motor$total_premium)

p_lr_age <- lr_by_age |>
  ggplot(aes(x = driver_age_band, y = agg_lr, group = 1)) +
  geom_ribbon(
    aes(ymin = agg_lr - 1.96 * se_lr,
        ymax = agg_lr + 1.96 * se_lr),
    fill = "#EF9A9A", alpha = 0.35
  ) +
  geom_line(linewidth = 1.3, colour = "#C62828") +
  geom_point(aes(size = n), colour = "#C62828") +
  geom_hline(yintercept = portfolio_lr,
             linetype = "dashed", colour = "grey40") +
  annotate("text",
           x     = 6, y = portfolio_lr,
           label = sprintf("Portfolio avg: %.1f%%", portfolio_lr * 100),
           vjust = -0.7, hjust = 1, size = 3.5, colour = "grey40") +
  scale_y_continuous(labels = percent) +
  scale_size_continuous(range = c(3, 9), labels = comma, name = "Policies") +
  labs(
    title    = "Aggregate Loss Ratio by Driver Age Band (95% CI ribbon)",
    subtitle = "Frequency is highest for young drivers, while aggregate LR is elevated in several non-young bands",
    x        = "Driver Age Band",
    y        = "Aggregate Loss Ratio"
  )

p_cv_age <- lr_by_age |>
  ggplot(aes(x = driver_age_band, y = cv_lr, group = 1)) +
  geom_line(linewidth = 1.2, colour = "#1565C0") +
  geom_point(size = 4, colour = "#1565C0") +
  geom_text(aes(label = round(cv_lr, 2)), vjust = -0.9, size = 3.5) +
  labs(
    title    = "Loss Ratio Volatility (Coefficient of Variation) by Driver Age",
    subtitle = "Higher CV signals less predictable outcomes and greater pricing uncertainty",
    x        = "Driver Age Band",
    y        = "Coefficient of Variation"
  )

p_lr_age / p_cv_age +
  plot_annotation(
    theme = theme(plot.title = element_blank())
  )
```

**Interpretation**: The driver-age signal is more nuanced than a pure U-shape. The **18-25 band** has the highest claim rate and exposure-adjusted claim frequency, making it a clear frequency-risk segment, but it does not have the highest aggregate loss ratio. Older bands have lower claim frequency but can still show elevated aggregate LR because severity and premium adequacy vary. The CV panel should therefore be read as volatility diagnostics, not as proof that the youngest band is always the least profitable.

#### 4.2.4 Loss Ratio by Geography

The same statistical comparison view is used for municipality type so geographic differences can be judged against the high dispersion in policy-level loss ratios.

```{r}
#| label: biv-lr-geo
#| fig-cap: "Municipality type significantly affects portfolio loss ratios"
#| fig-height: 8

set.seed(42)
ggbetweenstats(
  data                 = motor_lr |>
                           filter(!is.na(municipality_type)) |>
                           slice_sample(n = 50000),
  x                    = municipality_type,
  y                    = loss_ratio,
  type                 = "nonparametric",
  pairwise.comparisons = TRUE,
  p.adjust.method      = "holm",
  point.args           = list(alpha = 0.01, size = 0.3),
  title                = "Loss Ratio by Municipality Type",
  xlab                 = "Municipality Type",
  ylab                 = "Loss Ratio"
)
```

Circulation area has only two categories, so a two-group non-parametric comparison provides a direct rural-versus-urban profitability check.

```{r}
#| label: biv-lr-circulation
#| fig-cap: "Urban policyholders carry higher loss ratios than rural counterparts"
#| fig-height: 7

set.seed(42)
ggbetweenstats(
  data       = motor_lr |>
                 filter(!is.na(circulation_area)) |>
                 slice_sample(n = 50000),
  x          = circulation_area,
  y          = loss_ratio,
  type       = "nonparametric",
  point.args = list(alpha = 0.01, size = 0.3),
  title      = "Loss Ratio by Circulation Area",
  xlab       = "Circulation Area",
  ylab       = "Loss Ratio"
)
```

**Interpretation**:

-   Urban circulation areas show a modestly higher loss ratio and materially higher claim frequency than rural areas, consistent with denser traffic exposure. Municipality type is less clean: Coastal and Island loss ratios are close to Inland overall, and the Island pattern is better described as year-to-year volatility than persistently higher loss cost. Geography is still useful, but it should be credibility-weighted rather than converted directly into raw territorial relativities.

#### 4.2.5 Vehicle Value vs Loss Ratio

A scatterplot with marginal histograms is used to examine whether vehicle value relates to loss ratio while still showing the distribution of each variable separately.

```{r}
#| label: biv-vval-lr-marginal
#| fig-cap: "Vehicle value and loss ratio show a weak, non-linear relationship"
#| fig-height: 8

set.seed(42)
p_scatter <- motor_lr |>
  filter(!is.na(vehicle_value), vehicle_value > 0,
         vehicle_value < quantile(vehicle_value, 0.99, na.rm = TRUE),
         loss_ratio < 3) |>
  slice_sample(n = 20000) |>
  ggplot(aes(x = vehicle_value, y = loss_ratio,
             colour = policy_type)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_smooth(method = "loess", se = FALSE,
              colour = "black", linewidth = 1.2) +
  scale_x_continuous(labels = dollar_format(prefix = "EUR ")) +
  scale_y_continuous(labels = percent) +
  scale_colour_brewer(palette = "Set2", name = "Policy Type") +
  labs(
    title = "Vehicle Value vs Loss Ratio",
    x     = "Vehicle Value (EUR)",
    y     = "Loss Ratio"
  ) +
  theme(legend.position = "bottom")

ggMarginal(
  p_scatter,
  type   = "histogram",
  fill   = "#90CAF9",
  colour = "white",
  bins   = 40
)
```

**Interpretation**: Vehicle value should not be read as a simple linear driver of loss ratio in this portfolio. Higher-value vehicles do show higher claim frequency, but premiums and coverage mix offset part of the repair-cost effect, producing a weak and non-monotonic loss-ratio relationship. This makes vehicle value more useful as part of an interaction with policy type, deductible/excess, and repair-cost controls than as a standalone loading.

#### 4.2.6 Loss Ratio Distributions by Policy Type - Ridgeline Gradient

A ridgeline plot with a diverging colour gradient (blue = low LR, red = LR \> 1) exposes differences in the *shape* of loss ratio distributions across policy types. This matters for reserving: two segments with the same mean loss ratio can have very different tail weights, implying different capital requirements.

```{r}
#| label: ridgeline-lr-ptype
#| fig-cap: "Policy-type loss ratio distributions differ in both centre and tail shape"
#| fig-height: 7

motor_lr |>
  filter(loss_ratio < 5) |>
  ggplot(aes(x = loss_ratio, y = policy_type, fill = after_stat(x))) +
  geom_density_ridges_gradient(
    scale = 2, rel_min_height = 0.01,
    colour = "white", linewidth = 0.3
  ) +
  scale_fill_gradient2(
    low      = "#2196F3",
    mid      = "#FFF176",
    high     = "#E53935",
    midpoint = 1,
    name     = "Loss Ratio"
  ) +
  geom_vline(xintercept = 1,
             linetype = "dashed", colour = "black", linewidth = 0.8) +
  scale_x_continuous(labels = percent) +
  labs(
    title    = "Loss Ratio Distribution by Policy Type",
    subtitle = "Dashed line = break-even (LR = 1). Red shading = loss-making territory.",
    x        = "Loss Ratio",
    y        = "Policy Type"
  ) +
  theme_ridges(font_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

**Interpretation**:

-   All five policy types show their mass concentrated well below the 100% break-even line, confirming that claim-free policy-years dominate across the entire portfolio, consistent with the zero-inflation observed earlier. However, the tails tell a more important story. Every policy type extends into loss-making territory beyond 100%, with some reaching 400–500%, indicating that when losses do occur, they can be severe multiples of the premium collected. This is the core reinsurance design problem: the portfolio is profitable on average but exposed to individual policy-years where losses far exceed premium.

-   COMP_N shows the longest visible tail, extending furthest into red territory, which is consistent with its higher-value vehicle profile and greater claim severity potential. This segment represents the clearest case for reinsurance protection or conservative reserve loading. Critically, the near-identical shape of distributions across policy types, all sharply peaked near zero with similar tail behaviour, suggests that the key differentiator between segments is not the shape of the distribution but the frequency and magnitude of tail events. Segment-level profitability therefore hinges on how often these tail losses materialise, making claim frequency monitoring by policy type more actionable than average loss ratio comparisons alone.

### 4.3 Volatility Analysis - Claim Frequency and Severity

#### 4.3.1 Claim Severity by Policy Type - Statistical Validation

The raincloud plot earlier showed the shape of claim severity by policy type. Here the same comparison is validated statistically using a non-parametric test because severity is strongly right-skewed and heavy-tailed.

```{r}
#| label: biv-severity-ptype
#| fig-cap: "Claim severity differs significantly across policy types"
#| fig-height: 8

severity_sample <- motor |>
  filter(!is.na(claim_sev), claim_sev > 0,
         claim_sev < quantile(claim_sev, 0.975, na.rm = TRUE))

set.seed(42)
ggbetweenstats(
  data = severity_sample |> slice_sample(n = min(30000, nrow(severity_sample))),
  x = policy_type,
  y = claim_sev,
  type = "nonparametric",
  p.adjust.method = "holm",
  pairwise.comparisons = FALSE,
  point.args = list(alpha = 0.02, size = 0.4),
  title = "Claim Severity by Policy Type",
  xlab = "Policy Type",
  ylab = "Claim Severity (EUR per Claim)"
) +
  scale_y_continuous(labels = dollar_format(prefix = "EUR "))
```

**Interpretation**:

-   The statistical comparison supports the visual finding that claim cost distributions differ by coverage structure, but the direction is not simply "comprehensive equals higher severity." COMP_N has the highest claim frequency and aggregate LR, while several narrower covers show comparable or higher average severity among claim-bearing rows. Volatility management should therefore separate frequency control, premium adequacy, and large-loss reserving rather than treating severity as a single comprehensive-cover problem.

-   To make that volatility more operational, the next view translates ground-up policy-year incurred losses into illustrative excess layers. This is not treaty pricing, because the data does not include individual claim transactions, deductibles, limits, paid/outstanding splits, development, reinstatement terms, or expenses. It is still useful diagnostically: it shows which policy types feed the severe-loss layers that would matter most for retention and reinsurance decisions.

```{r}
#| label: large-loss-layer-summary

large_loss_layers <- motor |>
  filter(total_incurred > 0) |>
  mutate(
    layer_5k = pmax(total_incurred - 5000, 0),
    layer_10k = pmax(total_incurred - 10000, 0),
    layer_25k = pmax(total_incurred - 25000, 0),
    layer_50k = pmax(total_incurred - 50000, 0)
  ) |>
  group_by(policy_type) |>
  summarise(
    claim_rows = n(),
    incurred = sum(total_incurred),
    share_total_incurred = incurred / sum(motor$total_incurred),
    mean_claim_cost = mean(total_incurred),
    q99_claim_cost = quantile(total_incurred, 0.99),
    max_claim_cost = max(total_incurred),
    xs_10k_share = sum(layer_10k) / incurred,
    xs_25k_share = sum(layer_25k) / incurred,
    .groups = "drop"
  ) |>
  arrange(desc(share_total_incurred))

large_loss_layers |>
  transmute(
    `Policy Type` = policy_type,
    `Claim-Bearing Rows` = comma(claim_rows),
    `Share of Portfolio Incurred` = percent(share_total_incurred, 0.1),
    `Mean Claim Cost` = dollar(mean_claim_cost, prefix = "EUR ", accuracy = 1),
    `99th Percentile Claim Cost` = dollar(q99_claim_cost, prefix = "EUR ", accuracy = 1),
    `Maximum Claim Cost` = dollar(max_claim_cost, prefix = "EUR ", accuracy = 1),
    `Share Above EUR 10k Layer` = percent(xs_10k_share, 0.1),
    `Share Above EUR 25k Layer` = percent(xs_25k_share, 0.1)
  ) |>
  knitr::kable(caption = "Large-Loss Layer Diagnostics by Policy Type")
```

The grouped bar chart translates the layer table into a direct comparison of attachment-point sensitivity across policy types.

```{r}
#| label: large-loss-layer-plot
#| fig-cap: "Illustrative excess-layer burden by policy type"
#| fig-height: 7

large_loss_layers |>
  select(policy_type, xs_10k_share, xs_25k_share) |>
  pivot_longer(
    -policy_type,
    names_to = "Layer",
    values_to = "Share"
  ) |>
  mutate(
    Layer = recode(
      Layer,
      "xs_10k_share" = "Incurred above EUR 10k",
      "xs_25k_share" = "Incurred above EUR 25k"
    )
  ) |>
  ggplot(aes(x = fct_reorder(policy_type, Share), y = Share, fill = Layer)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("#1565C0", "#B71C1C"), name = NULL) +
  labs(
    title = "Which Policy Types Feed the Illustrative Excess Layers?",
    subtitle = "Higher shares above EUR 10k and EUR 25k indicate stronger large-loss sensitivity, not necessarily poor attritional pricing",
    x = NULL,
    y = "Share of Segment Incurred Loss Above Attachment"
  ) +
  theme(legend.position = "bottom")
```

**Reinsurance reading**:

-   The large-loss burden is not confined to comprehensive products. CC contributes the largest share of portfolio incurred loss because it is the dominant exposure class, COMP_E contains the maximum observed policy-year loss, and TPG shows a high proportion of its own incurred loss above the illustrative attachment points. If management’s objective is earnings smoothing, the data still points more naturally toward excess-of-loss protection than a pure quota share, but the treaty discussion would need claim-level bordereaux and development-adjusted ultimate losses before any pricing conclusion.

#### 4.3.2 Claim Frequency by Driver Age Band

Claim frequency (claims per exposure unit) is a direct measure of how accident-prone a segment is, independent of repair costs. Tracking both the raw claim rate and the exposure-adjusted frequency captures two complementary aspects of segment risk. Faceted line charts are used to show whether the two frequency measures move together across ordered age bands while allowing each metric to keep its own scale.

```{r}
#| label: biv-cf-age
#| fig-cap: "Young drivers generate disproportionately more claims relative to their exposure"
#| fig-height: 6

motor |>
  filter(!is.na(driver_age_band), total_exposure > 0) |>
  group_by(driver_age_band) |>
  summarise(
    claim_rate = mean(has_claim),
    mean_freq  = mean(claim_freq, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  pivot_longer(c(claim_rate, mean_freq),
               names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric,
    "claim_rate" = "Claim Rate (% Policies with Claims)",
    "mean_freq"  = "Mean Claim Frequency (Claims / Exposure)"
  )) |>
  ggplot(aes(x = driver_age_band, y = value,
             group = metric, colour = metric)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 3.5) +
  facet_wrap(~metric, scales = "free_y") +
  scale_y_continuous(labels = percent) +
  scale_colour_manual(
    values = c("#C62828", "#1565C0"), guide = "none"
  ) +
  labs(
    title = "Claim Frequency and Rate by Driver Age Band",
    x     = "Driver Age Band",
    y     = NULL
  ) +
  theme(strip.text = element_text(face = "bold"))
```

**Interpretation**:

-   The data confirms the well-established actuarial pattern of age-related claim frequency, but the magnitude here is striking. Drivers aged 18-25 show a claim rate above 20% and a mean claim frequency approaching 75% of exposure, more than double the rate of the next age band and nearly three times that of the oldest drivers.

-   From a pricing perspective, this gradient is clear and actionable: driver age should be a primary rating factor, and the 18-25 band warrants a material premium loading relative to the rest of the book. If this loading is absent or insufficient, the portfolio is cross-subsidising young drivers at the expense of older, lower-risk policyholders, a structural pricing problem that compounds over time as the mix of young drivers grows.

-   The near-monotonic decline across all age bands also means there is no evidence of a late-age frequency uptick, which is sometimes observed in very elderly drivers. The 65+ segment remains the lowest-frequency group, suggesting the portfolio does not currently face elevated risk from older drivers, though severity for this group should be examined separately before drawing conclusions about their overall profitability.

#### 4.3.3 Fair Comparison of Claim Frequency - Funnel Plot

A funnel plot is useful when segment sizes are unequal. Small segments naturally show more random variation, while large segments should sit closer to the portfolio average. The plot below compares policy type × bonus score × driver age cells against approximate 95% and 99.8% control limits for exposure-adjusted claim frequency.

```{r}
#| label: funnel-claim-frequency
#| fig-cap: "Exposure-adjusted claim frequency funnel plot by underwriting cell"
#| fig-height: 8

portfolio_freq <- sum(motor$total_claims, na.rm = TRUE) /
  sum(motor$total_exposure, na.rm = TRUE)

funnel_segments <- motor |>
  filter(!is.na(driver_age_band), !is.na(bonus_score), total_exposure > 0) |>
  group_by(policy_type, bonus_score, driver_age_band) |>
  summarise(
    policies = n(),
    exposure = sum(total_exposure, na.rm = TRUE),
    claims = sum(total_claims, na.rm = TRUE),
    claim_frequency = claims / exposure,
    .groups = "drop"
  ) |>
  filter(exposure >= 250) |>
  mutate(
    segment = paste(policy_type, bonus_score, driver_age_band, sep = " | "),
    upper_95 = portfolio_freq + qnorm(0.975) * sqrt(portfolio_freq / exposure),
    lower_95 = pmax(0, portfolio_freq - qnorm(0.975) * sqrt(portfolio_freq / exposure)),
    upper_998 = portfolio_freq + qnorm(0.999) * sqrt(portfolio_freq / exposure),
    lower_998 = pmax(0, portfolio_freq - qnorm(0.999) * sqrt(portfolio_freq / exposure)),
    status = case_when(
      claim_frequency > upper_998 ~ "High outlier",
      claim_frequency < lower_998 ~ "Low outlier",
      claim_frequency > upper_95 | claim_frequency < lower_95 ~ "Outside 95%",
      TRUE ~ "Within limits"
    )
  )

limit_grid <- tibble(
  exposure = seq(min(funnel_segments$exposure), max(funnel_segments$exposure), length.out = 300),
  upper_95 = portfolio_freq + qnorm(0.975) * sqrt(portfolio_freq / exposure),
  lower_95 = pmax(0, portfolio_freq - qnorm(0.975) * sqrt(portfolio_freq / exposure)),
  upper_998 = portfolio_freq + qnorm(0.999) * sqrt(portfolio_freq / exposure),
  lower_998 = pmax(0, portfolio_freq - qnorm(0.999) * sqrt(portfolio_freq / exposure))
)

ggplot(funnel_segments, aes(x = exposure, y = claim_frequency)) +
  geom_ribbon(
    data = limit_grid,
    aes(x = exposure, ymin = lower_998, ymax = upper_998),
    inherit.aes = FALSE,
    fill = "#FEE0B6", alpha = 0.55
  ) +
  geom_ribbon(
    data = limit_grid,
    aes(x = exposure, ymin = lower_95, ymax = upper_95),
    inherit.aes = FALSE,
    fill = "#B3CDE3", alpha = 0.65
  ) +
  geom_hline(yintercept = portfolio_freq, linetype = "dashed", colour = "grey30") +
  geom_point(aes(colour = status, size = policies), alpha = 0.85) +
  geom_text_repel(
    data = funnel_segments |> filter(status == "High outlier") |> slice_max(claim_frequency, n = 8),
    aes(label = segment),
    size = 3,
    max.overlaps = 20,
    box.padding = 0.35,
    seed = 42
  ) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = percent) +
  scale_size_continuous(range = c(2.5, 8), labels = comma, name = "Policies") +
  scale_colour_manual(
    values = c(
      "High outlier" = "#B71C1C",
      "Low outlier" = "#1565C0",
      "Outside 95%" = "#F9A825",
      "Within limits" = "grey55"
    ),
    name = "Segment Status"
  ) +
  labs(
    title = "Funnel Plot of Segment Claim Frequency",
    subtitle = "Shaded bands show expected random variation around the portfolio frequency; labelled cells are materially high-frequency outliers",
    x = "Total Exposure",
    y = "Claim Frequency"
  ) +
  theme(legend.position = "bottom")
```

**Interpretation**: The funnel plot separates true high-frequency cells from noise caused by unequal exposure. The most extreme outliers are not only young or poor-bonus cells: several COMP_N/Good cells also sit above the 99.8% limit, showing that coverage structure can dominate the claim-frequency signal after exposure is allowed for. Young-driver and Bad/Neutral bonus cells remain important, but the stronger operational lesson is to investigate each labelled cell's product design, deductible/excess structure, and claim-count definition before imposing a broad class surcharge.

#### 4.3.4 Bonus Score and Claim Occurrence - Chi-Square Test

I choose a stacked bar chart with an embedded chi-square test to assess whether the proportion of policies that generate at least one claim differs significantly across bonus score groups, directly testing the discriminatory power of the bonus-malus classification.

```{r}
#| label: barstats-bonus-claim
#| fig-cap: "Claim occurrence rates differ significantly across bonus score tiers"
#| fig-height: 7

set.seed(42)
ggbarstats(
  data         = motor |> slice_sample(n = 50000),
  x            = has_claim,
  y            = bonus_score,
  title        = "Claim Occurrence by Bonus Score",
  xlab         = "Bonus Score",
  legend.title = "Had a Claim"
)
```

#### 4.3.5 Correlation Structure of Numerical Risk Factors

A Spearman correlation matrix identifies which numerical variables move together across policies. In an insurance context, understanding whether driver age, vehicle value, premium, and loss ratio are correlated helps identify redundant rating factors and flags variables that may be proxying for each other.

```{r}
#| label: corrmat
#| fig-cap: "Spearman correlations among key numerical variables — x marks non-significant pairs"
#| fig-height: 9
#| fig-width: 12

set.seed(42)
motor |>
  filter(!is.na(loss_ratio), !is.na(vehicle_value),
         !is.na(driving_exp), driving_exp > 0) |>
  slice_sample(n = 30000) |>
  select(
    driver_age, vehicle_age, vehicle_value,
    power_to_weight_ratio, driving_exp,
    total_premium, total_exposure,
    loss_ratio, claim_freq, claim_sev
  ) |>
  ggcorrmat(
    type     = "nonparametric",
    colors   = c("#1565C0", "white", "#C62828"),
    title    = "Spearman Correlation Matrix of Key Numerical Variables",
    subtitle = "x marks non-significant correlations (alpha = 0.05)"
  )
```

**Interpretation**:

-   The numerical correlations are statistically useful but mostly modest in practical size, apart from the mechanical link between claim frequency and loss ratio. `loss_ratio` is weakly lower for older drivers and weakly higher for vehicle value in the sampled matrix, but neither relationship is strong enough to justify a standalone pricing rule. `claim_freq` and `claim_sev` are only weakly correlated with each other, supporting a frequency-severity framework rather than a single aggregate loss model.

#### 4.3.6 Multivariate Risk Profiling - Parallel Coordinates

Parallel coordinates allow patterns across six risk dimensions to be examined simultaneously without dimensionality reduction. Each line traces one policy from left to right across normalised axes. Colouring by bonus score reveals whether the "Bad" tier is systematically positioned in high-risk regions across multiple variables at once.

```{r}
#| label: parallel-coords
#| fig-cap: "Parallel coordinates reveal overlapping risk profiles by bonus score"
#| fig-height: 7

set.seed(42)
pc_data <- motor |>
  filter(!is.na(bonus_score), !is.na(vehicle_value),
         !is.na(loss_ratio), loss_ratio < 3,
         vehicle_value > 0, driving_exp > 0) |>
  slice_sample(n = 5000) |>
  select(bonus_score, driver_age, vehicle_age,
         driving_exp, vehicle_value, total_premium, loss_ratio)

ggparcoord(
  data        = pc_data,
  columns     = 2:7,
  groupColumn = 1,
  scale       = "uniminmax",
  alphaLines  = 0.2,
  title       = "Parallel Coordinates: Risk Factor Profiles by Bonus Score"
) +
  scale_colour_manual(
    values = c("Good" = "#1976D2", "Neutral" = "#FFA000", "Bad" = "#C62828"),
    name   = "Bonus Score"
  ) +
  scale_x_discrete(labels = c(
    "driver_age"    = "Driver Age",
    "vehicle_age"   = "Vehicle Age",
    "driving_exp"   = "Driving Exp.",
    "vehicle_value" = "Vehicle Value",
    "total_premium" = "Premium",
    "loss_ratio"    = "Loss Ratio"
  )) +
  labs(x = NULL, y = "Normalised Value (0-1)") +
  theme(
    axis.text.x     = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

**Interpretation**:

-   "Bad" bonus score policyholders (red) tend to appear more often in higher-risk regions of the panel, especially around younger/lower-experience profiles and higher loss-ratio paths. The plot is exploratory rather than causal, but it supports the view that bonus score is not an isolated label; it overlaps with other observable risk characteristics that should be checked together in pricing and underwriting review.

### 4.4 Temporal Analysis

#### 4.4.1 Portfolio Growth and Profitability Trends

The temporal view combines bars and line charts because policy count, loss ratio, and premium-versus-incurred movement answer different trend questions but need to be read together.

```{r}
#| label: temp-trends
#| fig-cap: "Portfolio expanded year-on-year while the aggregate loss ratio remained below break-even"
#| fig-height: 9

annual <- motor |>
  group_by(year) |>
  summarise(
    policies   = n(),
    total_prem = sum(total_premium),
    total_inc  = sum(total_incurred),
    agg_lr     = total_inc / total_prem,
    claim_rate = mean(has_claim),
    .groups    = "drop"
  )

p_growth <- annual |>
  ggplot(aes(x = year, y = policies, group = 1)) +
  geom_col(fill = "#1976D2", alpha = 0.85) +
  geom_text(aes(label = comma(policies)), vjust = -0.5, size = 4.5) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Policy Count by Year", x = NULL, y = "Policies")

p_lr_yr <- annual |>
  ggplot(aes(x = year, y = agg_lr, group = 1)) +
  geom_line(linewidth = 1.6, colour = "#C62828") +
  geom_point(size = 5, colour = "#C62828") +
  geom_text(aes(label = percent(agg_lr, 0.1)),
            vjust = -1.2, size = 4.5) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_continuous(labels = percent, limits = c(0, 1.5)) +
  labs(title = "Aggregate Loss Ratio by Year", x = NULL, y = "Agg. LR")

p_income <- annual |>
  ggplot(aes(x = year, group = 1)) +
  geom_line(aes(y = total_prem / 1e6, colour = "Premium"),
            linewidth = 1.4) +
  geom_line(aes(y = total_inc  / 1e6, colour = "Incurred"),
            linewidth = 1.4) +
  geom_point(aes(y = total_prem / 1e6, colour = "Premium"), size = 4) +
  geom_point(aes(y = total_inc  / 1e6, colour = "Incurred"), size = 4) +
  scale_colour_manual(
    values = c("Premium" = "#1565C0", "Incurred" = "#B71C1C")
  ) +
  scale_y_continuous(labels = \(x) paste0("EUR ", x, "M")) +
  labs(title = "Total Premium vs. Incurred Losses",
       x = NULL, y = NULL, colour = NULL) +
  theme(legend.position = "bottom")

(p_growth | p_lr_yr) / p_income
```

**Interpretation**:

-   The portfolio grew 150% over two years, from 67,172 to 168,133 policies, with premium and incurred losses scaling broadly in proportion. The book remains profitable, but the aggregate loss ratio has deteriorated from 65.6% to 74.7%, a 9-percentage-point rise that, while not alarming in isolation, narrows the buffer to break-even and warrants monitoring.

<!-- -->

-   The deterioration coincides with rapid growth, which is consistent with new business carrying higher loss propensity before seasoning. However, whether the 2024 uptick reflects a temporary new-business effect or a structural shift in claims experience cannot be determined from aggregate trends alone, segment and cohort-level analysis is needed to distinguish between the two.

#### 4.4.2 Loss Ratio Heatmap - Year × Segment

A tile heatmap displays aggregate loss ratio across two categorical dimensions simultaneously. The diverging colour scale (blue = low LR, red = high LR) immediately highlights which year-segment combinations are performing well or deteriorating, a view that would require many separate line charts to replicate.

```{r}
#| label: heatmap-lr
#| fig-cap: "Aggregate loss ratio across year and key categorical segments"
#| fig-height: 10

make_lr_heatmap <- function(seg_var, title) {
  motor |>
    filter(!is.na(!!sym(seg_var))) |>
    group_by(year, !!sym(seg_var)) |>
    summarise(
      agg_lr = sum(total_incurred) / sum(total_premium),
      .groups = "drop"
    ) |>
    ggplot(aes(x = year, y = !!sym(seg_var), fill = agg_lr)) +
    geom_tile(colour = "white", linewidth = 0.8) +
    geom_text(aes(label = percent(agg_lr, 0.1)),
              size = 4.5, fontface = "bold") +
    scale_fill_gradient2(
      low      = "#1565C0",
      mid      = "white",
      high     = "#C62828",
      midpoint = 0.5,
      name     = "Agg. LR",
      labels   = percent
    ) +
    labs(title = title, x = "Year", y = NULL) +
    theme(
      plot.title = element_text(face = "bold", size = 11),
      axis.text  = element_text(face = "bold"),
      panel.grid = element_blank()
    )
}

h1 <- make_lr_heatmap("policy_type",      "Year × Policy Type")
h2 <- make_lr_heatmap("bonus_score",      "Year × Bonus Score")
h3 <- make_lr_heatmap("municipality_type","Year × Municipality")
h4 <- make_lr_heatmap("circulation_area", "Year × Circulation Area")

(h1 | h2) / (h3 | h4) +
  plot_annotation(
    title = "Aggregate Loss Ratio Heatmaps — Year vs. Key Segments",
    theme = theme(plot.title = element_text(face = "bold", size = 15))
  )
```

**Interpretation**:

-   COMP_N is the most problematic policy type in the heatmap, with the highest aggregate LR in 2024 and a deterioration rather than a clean improvement. Bonus score also varies by year: Neutral is the worst cell in 2024, while Bad is adverse in 2023-2024 but not consistently above the portfolio average across all years. Island municipalities show the most erratic year-on-year movement, so the geographic story is more about volatility and credibility than a persistently higher Island loading.

------------------------------------------------------------------------

## 5. Additional Analysis:

### 5.1 Risk Flow Through the Portfolio

An alluvial (Sankey-style) diagram traces how policies flow across multiple categorical dimensions in a single visual. This would help identify the exact pathways through which claim volume concentrates in the portfolio.

```{r}
#| label: bonus-alluvial
#| fig-cap: "Policy flow from business type through bonus score to claim outcome"
#| fig-height: 9

set.seed(42)
motor |>
  slice_sample(n = 80000) |>
  filter(!is.na(bonus_score)) |>
  mutate(claim_outcome = if_else(has_claim, "Claim", "No Claim")) |>
  count(business_type, bonus_score, claim_outcome) |>
  ggplot(aes(
    axis1 = business_type,
    axis2 = bonus_score,
    axis3 = claim_outcome,
    y     = n
  )) +
  geom_alluvium(aes(fill = bonus_score), alpha = 0.65, width = 0.3) +
  geom_stratum(
    width = 0.3, fill = "grey88",
    colour = "white", linewidth = 0.5
  ) +
  geom_text(
    stat      = "stratum",
    aes(label = after_stat(stratum)),
    size = 4, fontface = "bold"
  ) +
  scale_fill_manual(
    values = c("Good" = "#1976D2", "Neutral" = "#FFA000", "Bad" = "#C62828"),
    name   = "Bonus Score"
  ) +
  scale_x_discrete(
    limits = c("Business Type", "Bonus Score", "Claim Outcome"),
    expand = c(0.05, 0.05)
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Policy Flow: Business Type → Bonus Score → Claim Outcome",
    y     = "Number of Policies"
  ) +
  theme(
    axis.text.x     = element_text(face = "bold", size = 12),
    legend.position = "bottom"
  )
```

**Interpretation**:

-   Good bonus-score policies dominate the no-claim stream because they dominate the portfolio by volume. The more critical insight is in the smaller flows: Bad and Neutral bonus tiers generate higher claim rates, and existing Portfolio business with Bad bonus has a particularly high claim occurrence rate. New Business Bad risks still contribute slightly more claim events than their policy share, but the alluvial chart should be read as a screening view; rate-based tables are needed before tightening new-business rules broadly.

### 5.2 Visual Model of Claim Propensity

The bivariate plots identify important risk factors one at a time. To check whether those signals remain visible after controlling for other factors, a logistic regression is fitted to claim occurrence and visualised as odds ratios. This is not intended as a production pricing model; it is a visual analytics check that helps separate overlapping factors.

```{r}
#| label: claim-model-data

set.seed(42)
claim_model_data <- motor |>
  filter(
    !is.na(driver_age_band),
    !is.na(bonus_score),
    !is.na(policy_type),
    !is.na(circulation_area),
    !is.na(municipality_type),
    !is.na(business_type)
  ) |>
  mutate(
    claim_flag = as.integer(has_claim),
    policy_type = fct_relevel(policy_type, "TP"),
    bonus_score_model = factor(as.character(bonus_score),
                               levels = c("Good", "Neutral", "Bad")),
    driver_age_band = fct_relevel(driver_age_band, "36-45"),
    circulation_area = fct_relevel(circulation_area, "Rural"),
    municipality_type = fct_relevel(municipality_type, "Inland"),
    business_type = fct_relevel(business_type, "Portfolio")
  ) |>
  (\(x) slice_sample(x, n = min(100000, nrow(x))))()

claim_glm <- glm(
  claim_flag ~ policy_type + bonus_score_model + driver_age_band +
    circulation_area + municipality_type + business_type,
  data = claim_model_data,
  family = binomial()
)

claim_or <- tidy(claim_glm, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    term_label = term |>
      str_replace("^policy_type", "Policy: ") |>
      str_replace("^bonus_score_model", "Bonus: ") |>
      str_replace("^driver_age_band", "Driver age: ") |>
      str_replace("^circulation_area", "Circulation: ") |>
      str_replace("^municipality_type", "Municipality: ") |>
      str_replace("^business_type", "Business: "),
    significant = if_else(conf.low > 1 | conf.high < 1, "Significant", "Not significant")
  )
```

An odds-ratio dot-and-interval plot is used because it communicates both direction and uncertainty after adjustment, while the log scale keeps protective and adverse effects visually comparable.

```{r}
#| label: claim-model-odds
#| fig-cap: "Adjusted odds ratios for claim occurrence"
#| fig-height: 8

claim_or |>
  ggplot(aes(x = estimate, y = fct_reorder(term_label, estimate), colour = significant)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.18, linewidth = 0.7) +
  geom_point(size = 3) +
  scale_x_log10(labels = number_format(accuracy = 0.01)) +
  scale_colour_manual(
    values = c("Significant" = "#B71C1C", "Not significant" = "grey60"),
    guide = "none"
  ) +
  labs(
    title = "Adjusted Claim Propensity Model",
    subtitle = "Odds ratios above 1 indicate higher claim odds after controlling for the other displayed factors",
    x = "Odds Ratio (log scale)",
    y = NULL,
    caption = "Reference categories: TP policy, Good bonus score, age 36-45, Rural, Inland, Portfolio business."
  )
```

**Interpretation**:

-   The model view confirms that several visual patterns remain after adjustment. Policy type is the strongest adjusted claim-propensity signal, especially COMP_N and COMP_E relative to TP. Poorer bonus history and younger driver age also continue to raise claim odds, while New Business has lower adjusted claim odds than Portfolio business in this sample. This strengthens the bivariate findings, but it should still be treated as a diagnostic model rather than a production pricing model.

## References

### Course References

-   *R for Visual Analytics*: [Chapter 1 - A Layered Grammar of Graphics: ggplot2 methods](https://r4va.netlify.app/chap01)
-   *R for Visual Analytics*: [Chapter 2 - Beyond ggplot2 Fundamentals](https://r4va.netlify.app/chap02)
-   *R for Visual Analytics*: [Chapter 3 - Programming Interactive Data Visualisation with R](https://r4va.netlify.app/chap03)
-   *R for Visual Analytics*: [Chapter 6 - Visual Correlation Analysis](https://r4va.netlify.app/chap06)
-   *R for Visual Analytics*: [Chapter 7 - Scatterplot with Marginal Distribution](https://r4va.netlify.app/chap07)
-   *R for Visual Analytics*: [Chapter 8 - Funnel Plots for Fair Comparisons](https://r4va.netlify.app/chap08)
-   *R for Visual Analytics*: [Chapter 9 - Visualising Distribution](https://r4va.netlify.app/chap09)
-   *R for Visual Analytics*: [Chapter 10 - Visual Statistical Analysis](https://r4va.netlify.app/chap10)
-   *R for Visual Analytics*: [Chapter 11 - Visualising Uncertainty](https://r4va.netlify.app/chap11)
-   *R for Visual Analytics*: [Chapter 14 - Heatmap for Visualising and Analysing Multivariate Data](https://r4va.netlify.app/chap14)
-   *R for Visual Analytics*: [Chapter 15 - Visual Multivariate Analysis with Parallel Coordinates Plot](https://r4va.netlify.app/chap15)
-   *R for Visual Analytics*: [Chapter 16 - Treemap Visualisation with R](https://r4va.netlify.app/chap16)
-   *R for Visual Analytics*: [Chapter 17 - Visualising and Analysing Time-oriented Data](https://r4va.netlify.app/chap17)
-   *R for Visual Analytics*: [Chapter 19 - Visualising, Analysing and Forecasting Time-series Data: tidyverts methods](https://r4va.netlify.app/chap19)
-   *R for Visual Analytics*: [Chapter 20 - Visualising, Analysing and Forecasting Time-series Data: timetk and modeltime methods](https://r4va.netlify.app/chap20)
-   *R for Visual Analytics*: [Chapter 31 - Information Dashboard Design: R methods](https://r4va.netlify.app/chap31)
-   *R for Visual Analytics*: [Chapter 33 - Summary](https://r4va.netlify.app/chap33)

### Method and Package References

-   [tidyverse](https://www.tidyverse.org/) for data import, tidying, transformation, grouping, and plotting workflow.
-   [ggplot2](https://ggplot2.tidyverse.org/) for layered grammar of graphics.
-   [ggridges](https://wilkelab.org/ggridges/) for ridgeline density plots.
-   [ggdist](https://mjskay.github.io/ggdist/) for distribution and uncertainty visualisation.
-   [ggstatsplot](https://indrajeetpatil.github.io/ggstatsplot/) for statistical graphics with embedded inferential results.
-   [ggExtra](https://cran.r-project.org/package=ggExtra) for scatterplots with marginal distributions.
-   [GGally](https://ggobi.github.io/ggally/) for correlation and parallel-coordinate visualisation.
-   [treemapify](https://wilkox.org/treemapify/) for treemap visualisation.
-   [ggalluvial](https://corybrunson.github.io/ggalluvial/) for alluvial flow visualisation.
-   [patchwork](https://patchwork.data-imaginist.com/) for composing multiple ggplot2 figures.
-   [ggrepel](https://ggrepel.slowkow.com/) for non-overlapping plot labels.
-   [broom](https://broom.tidymodels.org/) for converting model output into tidy tables for visualisation.

------------------------------------------------------------------------

## 6. Conclusion and Managerial Recommendation

| Theme | Key Finding |
|------------------------------------|------------------------------------|
| **Profitability** | COMP_N has the highest aggregate LR, with COMP_E also elevated; bonus-score effects are adverse but not perfectly ordered by aggregate LR |
| **Young drivers** | The 18-25 age band has the highest claim rate and frequency, but not the highest aggregate LR; it is primarily a frequency-risk segment |
| **Geography** | Urban circulation areas show higher claim frequency and modestly higher LR; Island municipalities are more volatile than persistently worse |
| **Vehicle value** | Higher vehicle value raises exposure to repair cost and claim frequency, but its relationship with LR is weak and non-linear after premium and coverage mix are considered |
| **Bonus-malus validation** | Chi-square, Kruskal-Wallis, and model diagnostics support bonus score as a useful risk signal, especially for claim occurrence and frequency |
| **Temporal** | The portfolio grew year-on-year; aggregate loss ratio remained below 100% across all three years, indicating sustained underwriting profitability |
| **Volatility** | Large-loss burden is distributed across several policy types: CC contributes the largest incurred share, COMP_E contains the maximum observed policy-year loss, and TPG has high excess-layer sensitivity |
| **Fair comparison** | Funnel analysis flags high-frequency cells that remain exceptional after accounting for unequal exposure size, including COMP_N/Good cells that would be missed by a simple young-driver or bad-bonus narrative |
| **Adjusted claim propensity** | The visual logistic model identifies policy type as the strongest adjusted claim-occurrence signal, while young drivers and poorer bonus scores remain material after controls |
| **Reinsurance lens** | The portfolio is profitable but tail-sensitive; excess-of-loss protection is conceptually more targeted than quota share, subject to claim-level treaty pricing data |

Effective risk segmentation across **policy type, driver age, bonus score, and geography** can improve underwriting discipline and pricing accuracy for this portfolio, but the relativities should be credibility-weighted rather than copied directly from raw segment loss ratios. The bonus-malus system is actuarially useful and should be preserved; young drivers warrant frequency-focused controls, while geography needs cautious territorial treatment because Island results are volatile rather than consistently worse. From a reinsurance perspective, the main concern is not aggregate unprofitability but **retained volatility**: severe policy-year losses are rare but financially meaningful, and the tail is spread across several material coverage classes.

| Observation | Decision implication |
|------------------------------------|------------------------------------|
| Aggregate LR is below 100%, but incurred losses are concentrated in the upper tail. | The book is profitable on average, yet retained earnings can still be volatile. Capital and reinsurance review should focus on tail load, not only mean LR. |
| COMP_N has the highest aggregate LR among policy types. | Comprehensive without excess may be underpriced for retained own-damage exposure, or may need stronger repair-network, vehicle-value, and excess/deductible controls. |
| Bad bonus and young-driver cells remain adverse after visual and model-based checks. | These are not just noisy segments; pricing relativities or underwriting rules should be reviewed for both intake and renewal decisions. |
| Urban and Island risks show higher or more erratic LR. | Geography may be proxying for repair inflation, traffic density, logistics cost, or sparse-volume volatility. A credibility-weighted territorial factor is preferable to simple raw LR ranking. |
| Frequency and severity are weakly coupled. | A frequency-severity view is required. One portfolio action may reduce accident count, while a different action is needed for severity leakage and large-loss retention. |

The reinsurance interpretation remains diagnostic rather than a treaty pricing conclusion. The dataset does not provide individual claim transactions, accident dates, claim development triangles, paid versus outstanding split, policy limits, deductibles, expenses, inflation trend, or reinstatement terms. A proper reinsurance submission would require historical ground-up loss bordereaux, exposure splits, rate-change history, large-loss narratives, and development-adjusted ultimate losses.
